---
title: 'Supplemental Materials for: Climatic Relationships to Agricultural Insurance Loss for the Pacific Northwest Region of the United States'
authors: "Erich Seamon, Paul E. Gessler, John T. Abatzoglou, Philip W. Mote, Stephen S. Lee"
date: "October 2023"
subtitle: 'The material contained herein is supplementary to the article named in the title and submitted to the Journal of Agriculture and Life Sciences.' 
always_allow_html: yes
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{setspace}\doublespacing
- \renewcommand{\figurename}{Figure S}
- \renewcommand{\tablename}{Table S}
output:
  pdf_document: default
  html_document:
    css: custom.css
    code_folding: hide
    highlight: textmate
    theme: united
    toc: true
    toc_float: true
params: null
mainfont: serif
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, fig.align = "center", out.width='400px', out.height='700px', dpi=400, dev = "cairo_pdf")
usePackage <- function(p) {
    if (!is.element(p, installed.packages()[,1]))
        install.packages(p, dep = TRUE, repos = "http://cran.us.r-project.org")
    require(p, character.only = TRUE)
}

usePackage("extrafont")
usePackage("ggplot2")
usePackage("webshot")
#webshot::install_phantomjs()
loadfonts()

```

This supplemental appendix provides exploratory data and factor analyses of agricultural insurance loss for the Pacific Northwest (PNW) and the inland Pacific Northwest (iPNW), to better understand the combined impacts by differing damage causes, commodities, counties, and years.


\newpage

## 1. Supplemental Tables List

Table S1. Example of insurance loss records that were acquired from the USDA Risk Management Agency (RMA).  Each record represents an individual insurance claim.  Full datasets are available via the following DOI: https://doi.org/10.5061/dryad.hhmgqnknh
<br></br>


\newpage

## 2. Supplemental Figures List

### Pacific Northwest (PNW) Summary Supplemental Figures

Figure S1. Pacific Northwest study area, which includes agricultural regions for the inland Pacific Northwest, the southern Idaho valley, and the Willamette valley.
<br></br>

Figure S2. PNW agricultural insurance loss by year, 1989 to 2022.
<br></br>

Figure S3. PNW agricultural insurance loss by year: 2001 to 2022.
<br></br>

Figure S4. PNW agricultural insurance loss by damage cause: 2001 to 2022.
<br></br>

Figure S5. PNW agricultural insurance loss by commodity: 2001 to 2022
<br></br>

### Inland Pacific Northwest (iPNW) Summary Supplemental Figures

Figure S6. iPNW agricultural insurance loss by year, 2001 to 2022.
<br></br>

Figure S7. iPNW agricultural insurance loss by damage cause, 2001 to 2022.
<br></br>

Figure S8. iPNW agricultural insurance loss by commodity, 2001 to 2022.
<br></br>

Figure S9. iPNW agricultural insurance loss for wheat by damage cause: 2001 to 2022.
<br></br>

Figure S10. iPNW stacked barplot of annual wheat agricultural insurance loss, by top damage causes, 2001 to 2022.
<br></br>

Figure S11. iPNW map of wheat losses, 2001 to 2022.
<br></br>

Figure S12. iPNW agricultural insurance loss for apples by damage cause: 2001 to 2022.
<br></br>

Figure S13. iPNW stacked barplot of annual apples agricultural insurance loss, by top damage causes, 2001 to 2022.
<br></br>

Figure S14. iPNW map of apple losses, 2001 to 2022.
<br></br>

### Pacific Northwest (iPNW) PCA Supplemental Figures

Figure S15. Top panel: biplot of principal components for PNW insurance loss, for wheat by county, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used.
<br></br>

Figure S16. Top panel: biplot of principal components for PNW insurance loss, for apples by county, with damage caus eas the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used.
<br></br>

Figure S17. Top panel: biplot of principal components for PNW insurance loss, for barley by county, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used.
<br></br>

### Inland Pacific Northwest (iPNW) PCA Supplemental Figures

Figure S18. Top panel: biplot of principal components for iPNW insurance loss, for all commodities by county, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used

Figure S19. Top panel: biplot of principal components for iPNW insurance loss, for all commodities by month, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used
<br></br>

Figure S20. Top panel: biplot of principal components for iPNW insurance loss, for wheat by year, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used
<br></br>

Figure S21. Top panel: biplot of principal components for iPNW insurance loss, for wheat by county, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used
<br></br>

Figure S22. Top panel: biplot of principal components for iPNW insurance loss, for barley by county, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used
<br></br>

Figure S23. Top panel: biplot of insurance loss for the iPNW, for wheat by year, with damage cause as the factor loadings.  Uses a Kmeans technique for grouping. Bottom panel: Scree plot. Data from 2001 is 2022 is used.
<br></br>

Figure S24. Top panel: biplot of insurance loss for the iPNW, for wheat by county, with damage cause as the factor loadings.  Uses a Kmeans technique for grouping. Bottom panel: Scree plot. Data from 2001 is 2022 is used.
<br></br>


```{r message=FALSE, warning=FALSE, echo=FALSE}

usePackage("knitr")
usePackage("tab")
usePackage("car")
usePackage("RCurl")
usePackage("lme4")
usePackage("ez")
usePackage("lattice")
usePackage("ggplot2")
usePackage("coefplot2")
usePackage("broom")
usePackage("htmlTable")
usePackage("gridExtra")
usePackage("kableExtra")
usePackage("repmis")



#options(scipen=999)


#------add crop insurance data

destfile <- "./data/RMA_originaldata/RMA_originaldata.zip"
#download.file(destfile, destfile)
outDir<-"/tmp/aginsurance/RMA_originaldata/"
unzip(destfile,exdir=outDir)


rma1989 <- read.csv("/tmp/aginsurance/RMA_originaldata/1989.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma1990 <- read.csv("/tmp/aginsurance/RMA_originaldata/1990.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma1991 <- read.csv("/tmp/aginsurance/RMA_originaldata/1991.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma1992 <- read.csv("/tmp/aginsurance/RMA_originaldata/1992.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma1993 <- read.csv("/tmp/aginsurance/RMA_originaldata/1993.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma1994 <- read.csv("/tmp/aginsurance/RMA_originaldata/1994.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma1995 <- read.csv("/tmp/aginsurance/RMA_originaldata/1995.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma1996 <- read.csv("/tmp/aginsurance/RMA_originaldata/1996.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma1997 <- read.csv("/tmp/aginsurance/RMA_originaldata/1997.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma1998 <- read.csv("/tmp/aginsurance/RMA_originaldata/1998.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma1999 <- read.csv("/tmp/aginsurance/RMA_originaldata/1999.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2000 <- read.csv("/tmp/aginsurance/RMA_originaldata/2000.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2001 <- read.csv("/tmp/aginsurance/RMA_originaldata/2001.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2002 <- read.csv("/tmp/aginsurance/RMA_originaldata/2002.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2003 <- read.csv("/tmp/aginsurance/RMA_originaldata/2003.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2004 <- read.csv("/tmp/aginsurance/RMA_originaldata/2004.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2005 <- read.csv("/tmp/aginsurance/RMA_originaldata/2005.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2006 <- read.csv("/tmp/aginsurance/RMA_originaldata/2006.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2007 <- read.csv("/tmp/aginsurance/RMA_originaldata/2007.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2008 <- read.csv("/tmp/aginsurance/RMA_originaldata/2008.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2009 <- read.csv("/tmp/aginsurance/RMA_originaldata/2009.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2010 <- read.csv("/tmp/aginsurance/RMA_originaldata/2010.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2011 <- read.csv("/tmp/aginsurance/RMA_originaldata/2011.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2012 <- read.csv("/tmp/aginsurance/RMA_originaldata/2012.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2013 <- read.csv("/tmp/aginsurance/RMA_originaldata/2013.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2014 <- read.csv("/tmp/aginsurance/RMA_originaldata/2014.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2015 <- read.csv("/tmp/aginsurance/RMA_originaldata/2015.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2016 <- read.csv("/tmp/aginsurance/RMA_originaldata/2016.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2017 <- read.csv("/tmp/aginsurance/RMA_originaldata/2017.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2018 <- read.csv("/tmp/aginsurance/RMA_originaldata/2018.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2019 <- read.csv("/tmp/aginsurance/RMA_originaldata/2019.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2020 <- read.csv("/tmp/aginsurance/RMA_originaldata/2020.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2021 <- read.csv("/tmp/aginsurance/RMA_originaldata/2021.txt", sep = "|", header = FALSE, strip.white = TRUE)
rma2022 <- read.csv("/tmp/aginsurance/RMA_originaldata/2022.txt", sep = "|", header = FALSE, strip.white = TRUE)



#load individual files from 1989 to 2000
RMA_1989 <- rbind(rma1989, rma1990, rma1991, rma1992, rma1993, rma1994, rma1995, rma1996, rma1997, rma1998, rma1999, rma2000)

#load individual files from 2001 to 2015
RMA_2015 <- rbind(rma2001, rma2002, rma2003, rma2004, rma2005, rma2006, rma2007, rma2008, rma2009, rma2010, rma2011, rma2012, rma2013, rma2014, rma2015)

#load individual files from 2015 to 2023
RMA_2022 <- rbind(rma2016, rma2017, rma2018, rma2019, rma2020, rma2021, rma2022)

#filter columns
RMA_1989 <- data.frame(RMA_1989[,1], RMA_1989[,3], RMA_1989[,5], RMA_1989[,7], RMA_1989[,12], RMA_1989[,13], RMA_1989[,14], RMA_1989[,15])

#filter columns
RMA_2015 <- data.frame(RMA_2015[,1], RMA_2015[,3], RMA_2015[,5], RMA_2015[,7], RMA_2015[,12], RMA_2015[,13], RMA_2015[,14], RMA_2015[,15], RMA_2015[,16])

#filter columns
RMA_2022 <- data.frame(RMA_2022[,1], RMA_2022[,3], RMA_2022[,5], RMA_2022[,7], RMA_2022[,13], RMA_2022[,14], RMA_2022[,15], RMA_2022[,20], RMA_2022[,21])

#set column names
colnames(RMA_1989) <- c("year", "state", "county", "commodity", "damagecause", "monthcode", "month", "loss")

colnames(RMA_2015) <- c("year", "state", "county", "commodity", "damagecause", "monthcode", "month", "acres", "loss")

colnames(RMA_2022) <- c("year", "state", "county", "commodity", "damagecause", "monthcode", "month", "acres", "loss")

#load individual files from 2001 to 2015

RMA <- rbind(RMA_2015, RMA_2022)
#RMA <- rbind(rma2001, rma2002, rma2003, rma2004, rma2005, rma2006, rma2007, rma2008, rma2009, rma2010, rma2011, rma2012, rma2013, rma2014, rma2015, rma2016, rma2017, rma2018, rma2019, rma2020, rma2021, rma2022, rma2023 )

#filter columns
#RMA <- data.frame(RMA[,1], RMA[,3], RMA[,5], RMA[,7], RMA[,12], RMA[,13], RMA[,14], RMA[,15], RMA[,16])

#set column names
#colnames(RMA) <- c("year", "state", "county", "commodity", "damagecause", "monthcode", "month", "acres", "loss")

#subset 2001 to 2015 for ID, WA, and OR
RMA_PNW <- subset(RMA, state == "WA" | state == "OR" | state == "ID" )

#subset 1989 to 2000 for ID, WA, and OR
RMA_PNW_1989 <- subset(RMA_1989, state == "WA" | state == "OR" | state == "ID" )

#calculate loss per acre
RMA_PNW$lossperacre <- RMA_PNW$loss / RMA_PNW$acres

#remove records with missing months
RMA_PNW <- subset(RMA_PNW, month != "")

#set a crop year column (i.e. water year)
RMA_PNW$cropyear <- RMA_PNW$year

#make oct/nov/dec part of the crop year
RMA_PNW$cropyear[RMA_PNW$monthcode >= 10] <- RMA_PNW$year + 1

#remove 2016
#RMA_PNW <- subset(RMA_PNW, year != 2016)





#subset just for Idaho
RMA_Idaho_PNW <- subset(RMA_PNW, state == "ID")

#subset just for Oregon
RMA_Oregon_PNW <- subset(RMA_PNW, state == "OR")

#subset just for Washington
RMA_Washington_PNW <- subset(RMA_PNW, state == "WA")

#subset just for iPNW counties for Idaho
RMA_Idaho_IPNW <- subset(RMA_Idaho_PNW, county == "Idaho" | county == "Lewis" | county ==  "Nez Perce" | county ==  "Latah" | county ==  "Benewah")

#subset just for iPNW counties for Oregon
RMA_Oregon_IPNW <- subset(RMA_Oregon_PNW, county == "Wasco" | county ==  "Sherman" | county ==  "Gilliam" | county ==  "Morrow" | county ==  "Umatilla" | county ==  "Union" | county ==  "Wallowa")

#subset just for iPNW counties for Washington
RMA_Washington_IPNW <- subset(RMA_Washington_PNW, county ==  "Douglas" | county ==  "Grant" | county ==  "Benton" | county ==  "Franklin" | county ==  "Walla Walla" | county ==  "Adams" | county ==  "Lincoln" | county ==  "Spokane" | county ==  "Whitman" | county ==  "Columbia" | county ==  "Garfield" | county ==  "Asotin")


#combine all iPNW counties
RMA_IPNW <- rbind(RMA_Idaho_IPNW,RMA_Oregon_IPNW,RMA_Washington_IPNW)


#aggregate by year/state/county/commodity, for total loss in $
RMA_commodity_loss <- aggregate(RMA$loss, by = list(RMA$year, RMA$state, RMA$county, RMA$commodity), FUN = sum)
colnames(RMA_commodity_loss) <- c("year", "state", "county", "commodity", "loss")
RMA_commodity_loss <- subset(RMA_commodity_loss, commodity != "ADJUSTED GROSS REVENUE")

#aggregate by year/state/county/commodity for total number of claims
RMA_commodity_count <- aggregate(RMA$loss, by = list(RMA$year, RMA$state, RMA$county, RMA$commodity), FUN = length)
colnames(RMA_commodity_count) <- c("year", "state", "county", "commodity", "count")
RMA_commodity_count <- subset(RMA_commodity_count, commodity != "ADJUSTED GROSS REVENUE")

#aggregate for year/state/county/commodity for acreage
RMA_commodity_acres <- aggregate(RMA$acres, by = list(RMA$year, RMA$state, RMA$county, RMA$commodity), FUN = sum)
colnames(RMA_commodity_acres) <- c("year", "state", "county", "commodity", "acres")
RMA_commodity_acres <- subset(RMA_commodity_acres, commodity != "ADJUSTED GROSS REVENUE")

#aggregate for year/state/county/damage cause for total loss in $
RMA_damage_loss <- aggregate(RMA$loss, by = list(RMA$year, RMA$state, RMA$county, RMA$commodity, RMA$damagecause), FUN = sum)
colnames(RMA_damage_loss) <- c("year", "state", "county", "commodity", "damagecause", "loss")

#aggregate for year/state/county/commodity/damage cause by total loss in $
RMA_damage_count <- aggregate(RMA$loss, by = list(RMA$year, RMA$state, RMA$county, RMA$commodity, RMA$damagecause), FUN = length)
colnames(RMA_damage_count) <- c("year", "state", "county", "commodity", "damagecause", "count")

#aggregate by year/state/county/commodity/damage cause by acres
RMA_damage_acres <- aggregate(RMA$acres, by = list(RMA$year, RMA$state, RMA$county, RMA$commodity, RMA$damagecause), FUN = sum)
colnames(RMA_damage_acres) <- c("year", "state", "county", "commodity", "damaecause", "acres")

#combine all the aggregated commodity values (loss, count, acreage)
RMA_commodity_combined <- cbind(RMA_commodity_loss, RMA_commodity_count$count, RMA_commodity_acres$acres)
colnames(RMA_commodity_combined) <- c("year", "state", "county", "commodity", "loss", "count", "acres")

#calculate lossperacre for aggregated commodity
RMA_commodity_combined$lossperacre <- RMA_commodity_combined$loss / RMA_commodity_combined$acres

#calculate loss per claim for aggregated commodity
RMA_commodity_combined$lossperclaim <- RMA_commodity_combined$loss / RMA_commodity_combined$count

#calculate acres per claim for aggregated commodity
RMA_commodity_combined$acresperclaim <- RMA_commodity_combined$acres / RMA_commodity_combined$count

#combine all aggregated damage cause values (loss, count, acreage) 
RMA_damage_combined <- cbind(RMA_damage_loss, RMA_damage_count$count, RMA_damage_acres$acres)
colnames(RMA_damage_combined) <- c("year", "state", "county", "commodity", "damagecause", "loss", "count", "acres")

#calculate lossperacre for aggregated damage cause
RMA_damage_combined$lossperacre <- RMA_damage_combined$loss / RMA_damage_combined$acres

#calculate loss per claim for aggregated damage cause
RMA_damage_combined$lossperclaim <- RMA_damage_combined$loss / RMA_damage_combined$count

#calculate acres per claim for aggregated damage cause
RMA_damage_combined$acresperclaim <- RMA_damage_combined$acres / RMA_damage_combined$count

#set NAs for loss per acre for damage cause combo to zero.  
RMA_damage_combined$lossperacre[!is.finite(RMA_damage_combined$lossperacre)] <- 0

#subset for PNW - 2001 - 2015
RMA_damage_PNW <- subset(RMA_damage_combined, state == "WA" | state == "OR" | state == "ID")

#remove all other counties values
RMA_damage_PNW <- subset(RMA_damage_PNW, county != "All Other Counties")

#subset just for Idaho
RMA_Idaho <- subset(RMA_damage_PNW, state == "ID")

#subset just for Oregon
RMA_Oregon <- subset(RMA_damage_PNW, state == "OR")

#subset just for Washington
RMA_Washington <- subset(RMA_damage_PNW, state == "WA")

#subset just for iPNW counties for Idaho
RMA_Idaho_IPNW <- subset(RMA_Idaho, county == "Idaho" | county == "Lewis" | county ==  "Nez Perce" | county ==  "Latah" | county ==  "Benewah")

#subset just for iPNW counties for Oregon
RMA_Oregon_IPNW <- subset(RMA_Oregon, county == "Wasco" | county ==  "Sherman" | county ==  "Gilliam" | county ==  "Morrow" | county ==  "Umatilla" | county ==  "Union" | county ==  "Wallowa")

#subset just for iPNW counties for Washington
RMA_Washington_IPNW <- subset(RMA_Washington, county ==  "Douglas" | county ==  "Grant" | county ==  "Benton" | county ==  "Franklin" | county ==  "Walla Walla" | county ==  "Adams" | county ==  "Lincoln" | county ==  "Spokane" | county ==  "Whitman" | county ==  "Columbia" | county ==  "Garfield" | county ==  "Asotin")


#combine all iPNW counties
RMA_damage_IPNW <- rbind(RMA_Idaho_IPNW,RMA_Oregon_IPNW,RMA_Washington_IPNW)

 palousecounties    <-  c("Idaho", "Lewis", "Nez Perce", "Clearwater", "Latah", "Benewah", "Kootenai","Douglas", "Grant", "Benton", "Franklin", "Walla Walla", "Adams", "Lincoln", "Spokane", "Whitman", "Columbia", "Garfield", "Asotin","Wasco", "Sherman", "Gilliam", "Morrow", "Umatilla", "Union", "Wallowa")
 
sumloss_palouse <- RMA_damage_PNW[RMA_damage_PNW$county %in% palousecounties, ] 
sumloss_nonpalouse <- RMA_damage_PNW[!RMA_damage_PNW$county %in% palousecounties, ]

#----


```


```{r functions, message=FALSE, warning=FALSE, error=TRUE, echo = FALSE, include=FALSE}

# A function for captioning and referencing images

fig <- local({
    i <- 0
    ref <- list()
    list(
        cap=function(refName, text) {
            i <<- i + 1
            ref[[refName]] <<- i
            paste("Figure ", i, ": ", text, sep="")
        },
        ref=function(refName) {
            ref[[refName]]
        })
})
``` 

\newpage

```{r message=FALSE, warning=FALSE, echo = FALSE, fig.width = 8, fig.height = 7, fig.cap=paste("Pacific Northwest study area, which includes agricultural regions for the inland Pacific Northwest, the southern Idaho valley, and the Willamette valley.")}

usePackage("data.table")
usePackage("maptools")
usePackage("classInt")
usePackage("leaflet")
usePackage("dplyr")
usePackage("raster")
usePackage("htmltools")

destfile <- "./data/states/states_conus.zip"
#download.file(URL, destfile)
outDir<-"/tmp/aginsurance/states_conus/"
unzip(destfile,exdir=outDir)

states <- readShapePoly('/tmp/aginsurance/states_conus/states_conus.shp',
                        proj4string=CRS
                        ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")


destfile <- "./data/counties/threestate_boundary.zip"
#download.file(URL, destfile)
outDir<-"/tmp/aginsurance/threestate_boundary/"
unzip(destfile,exdir=outDir)
    
threestates <- readShapePoly('/tmp/aginsurance/threestate_boundary/threestate_boundary.shp', 
                                  proj4string=CRS
                                  ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
        projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")


#setwd("/dmine/data/counties/")
        
destfile <- "./data/counties/threestate_willamette.zip"
#download.file(URL, destfile)
outDir<-"/tmp/aginsurance/threestate_willamette/"
unzip(destfile,exdir=outDir)

counties_willamette <- readShapePoly('/tmp/aginsurance/threestate_willamette/threestate_willamette.shp',
                     proj4string=CRS
                     ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

destfile <- "./data/counties/threestate_southernID.zip"
#download.file(URL, destfile)
outDir<-"/tmp/aginsurance/threestate_southernID/"
unzip(destfile,exdir=outDir)

counties_ID <- readShapePoly('/tmp/aginsurance/threestate_southernID/threestate_southernID.shp',
                     proj4string=CRS
                     ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

destfile <- "./data/counties/threestate_palouse.zip"
#download.file(URL, destfile)
outDir<-"/tmp/aginsurance/threestate_palouse/"
unzip(destfile,exdir=outDir)

counties_palouse <- readShapePoly('/tmp/aginsurance/threestate_palouse/threestate_palouse.shp',
                     proj4string=CRS
                     ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

#pal <- colorNumeric(palette = c("white", "orange", "darkorange", "red", "darkred"),
#                    domain = counties$NAME)




exte <- as.vector(extent(threestates))

#label <- paste(sep = "<br/>", counties$NAME, round(counties$NAME, 0))
#markers <- data.frame(label)


#labs <- as.list(counties$NAME)

counties_palouse <- counties_palouse[counties_palouse$NAME != "Kootenai",]
counties_palouse <- counties_palouse[counties_palouse$NAME != "Clearwater",]

#--

par(mar=c(5, 8.1, 7, 0), family = 'serif', mgp=c(1, 1, 0), las=0)


usePackage("RgoogleMaps")
usePackage("sp")
usePackage("ggplot2")
usePackage("ggmap")
usePackage("maps")
usePackage("mapdata")


usePackage("mapproj")

maps::map("worldHires","Canada", xlim=c(-130,-110),ylim=c(40,50), col="gray90", fill=TRUE)  #plot the region of Canada I want

maps::map("worldHires","usa", xlim=c(-130,-110),ylim=c(40,50), col="gray95", fill=TRUE, add=TRUE)  #add the adjacent parts of the US; can't forget my homeland

plot(counties_palouse, add=TRUE, col=alpha("darkgreen", 0.6), border=FALSE)  
lines(counties_palouse)
lines(states)
box(lty = 1, col = 'black')


plot(counties_willamette, add=TRUE, col=alpha("red", 0.6), border=FALSE)  
lines(counties_willamette)


plot(counties_ID, add=TRUE, col=alpha("yellow", 0.6), border=FALSE)  
lines(counties_ID)

map.scale(x=-120.5, y=41, ratio=FALSE, relwidth=0.17) 
#north.arrow(xb = -121, yb=41, len = 0.22, lab="N")

# Inmap
par(usr=c(-129, 110, 22, 144))
rect(xleft =-126.2,ybottom = 23.8,xright = -65.5,ytop = 50.6,col = "white")
rect(xleft =60,ybottom = 75,xright = -35,ytop = 125,border = "red", col = "NA", lwd=2)

maps::map("usa", xlim=c(-126.2,-65.5), ylim=c(23.8,50.6),add=T)
maps::map("state", xlim=c(-126.2,-65.5), ylim=c(23.8,50.6),add=T, boundary = F, interior = T, lty=2)
#map("state", region="california", fill=T, add=T)
#points(-121.6945, 39.36708, bg = "white", pch = 21)
maps::map("state", region=c("washington", "Idaho", "Oregon"), fill=T, add=T) 

#mapz <- leaflet(options = leafletOptions(zoomSnap = .4, zoomDelta = .4)) %>%  
#  addProviderTiles(providers$Hydda.Base) %>%
 #  addProviderTiles(providers$Stamen.TonerLines) %>% 
 
#  fitBounds(exte[1], exte[3], exte[2], exte[4]) %>% addPolygons(data = counties_palouse, color = "lightyellow",  fillOpacity = .8, weight = 1) %>% addPolygons(data = counties_palouse, color = "gray",  fillOpacity = 0, weight = 1) %>% addPolygons(data = counties_ID, color = "gray", fillOpacity = 0,  weight = 1) %>% addPolygons(data = counties_ID, color = "red",  weight = 1) %>% addPolygons(data = counties_willamette, color = "green",  weight = 1)  %>% addPolygons(data = counties_willamette, color = "gray", fillOpacity = 0, weight = 1)  %>% addPolygons(data = states, stroke = TRUE, color = "black",  weight = 3, fillOpacity = 0) 

#addScaleBar(mapz, position = c("bottomright"), options = scaleBarOptions())

```

\newpage
\blandscape


```{r message=FALSE, warning=FALSE, echo=FALSE}

#```{r message=FALSE, warning=FALSE, error=TRUE, echo = FALSE, fig.width = 8, fig.height = 8, fig.cap=paste("Example of the original data that was acquired from the USDA Risk Management Agency (RMA).  Each record represents an individual insurance claim.")}

RMA_PNW_table <- RMA_PNW[-6]

knitr::kable(RMA_PNW_table[1:9,], row.names = FALSE, booktabs = T, format="latex", caption = "Example of insurance loss records that were acquired from the USDA Risk Management Agency (RMA).  Each record represents an individual insurance claim.  Full datasets are available via the following DOI: https://doi.org/10.5061/dryad.hhmgqnknh") %>%
   kable_styling(latex_options = c("hold_position"),
                full_width = F)

```




\newpage
\elandscape

```{r message=FALSE, warning=FALSE, error=TRUE, echo = FALSE, fig.width = 8, fig.height = 8, fig.cap=paste("Pacific Northwest agricultural insurance loss by year, 1989 to 2022")}

usePackage("htmlTable")
usePackage("gridExtra")

#palouse_sumloss_aggregate <- subset(palouse_sumloss_aggregate, year >= 2001 & year <= 2015)

#palouse_sumloss_1989_2015 <- subset(palouse_sumloss_aggregate, commodity == "WHEAT")
RMA_PNW2 <- RMA_PNW[-8]
RMA_PNW2 <- RMA_PNW2[-9]
RMA_PNW2 <- RMA_PNW2[-9]

RMA_all <- rbind(RMA_PNW_1989, RMA_PNW2)


#barplot of 1989 to 2015

RMA_PNW_aggregates <- aggregate(RMA_PNW$loss, by=list(RMA_PNW$year), FUN = "sum")
RMA_PNW_1989_aggregates <- aggregate(RMA_PNW_1989$loss, by=list(RMA_PNW_1989$year), FUN = "sum")
RMA_allyear <- rbind(RMA_PNW_1989_aggregates, RMA_PNW_aggregates)
colnames(RMA_allyear) <- c("Year", "Loss")

#barplot of 1989 to 2015 for damage causes

RMA_PNW_aggregates_d <- aggregate(RMA_PNW$loss, by=list(RMA_PNW$damagecause), FUN = "sum")
RMA_PNW_1989_aggregates_d <- aggregate(RMA_PNW_1989$loss, by=list(RMA_PNW_1989$damagecause), FUN = "sum")
RMA_allyear_d <- rbind(RMA_PNW_1989_aggregates_d, RMA_PNW_aggregates_d)
colnames(RMA_allyear_d) <- c("Damagecause", "Loss")

RMA_allyear_d <- aggregate(RMA_allyear_d$Loss, list(RMA_allyear_d$Damagecause), FUN = "sum")
colnames(RMA_allyear_d) <- c("Damagecause", "Loss")
RMA_allyear_d <- RMA_allyear_d[order(-RMA_allyear_d$Loss),] 

par(mar=c(5, 8.1, 7, 0), family = 'serif', mgp=c(1, 1, 0), las=0)
#levels(YYY1$commodity)[1] <- "ADJ GROSS REVENUE"
barplot(RMA_allyear$Loss, names.arg = RMA_allyear$Year, las = 3, main = "", yaxt="n", col = "lightblue", xlab = "", cex.main = 1, cex.lab = 1, cex.names = 1, font.lab = 1, font.axis = 1) 
text(x=24, y=1500000000, "83% of \n Agriculture claims occur \n after 2000", cex = 1.5)


graphics::title(main = "PNW total insurance loss by year: 1989 to 2022", line = 1, adj = 0, cex.main = 1.5)
pts <- pretty(RMA_allyear$Loss / 1000000)
pts2 <- pretty(RMA_allyear$Loss)
axis(2, las = 1, cex.axis = 1, at = pts2, labels = paste(pts, "M", sep = ""), cex.axis=1.5)
abline(v = 14.5, lty = 2, color = "red")

mtext("Year", cex = 1.5, side=1, line=4)
mtext("Insurance Claim Loss ($)", cex = 1.5, side=2, line=5.5)


```

\newpage

```{r message=FALSE, warning=FALSE, error=TRUE, echo = FALSE, fig.width = 8, fig.height = 8, fig.cap=paste("Pacific Northwest agricultural insurance loss by year: 2001 to 2022")}

palouse_sumloss_1989_2015 <- aggregate(RMA_PNW$loss, list(RMA_PNW$year), FUN = "sum")

X <- palouse_sumloss_1989_2015
colnames(X) <- c("year", "loss")
#X <- palouse_sumloss_2007_2015_aggregate[order(palouse_sumloss_2007_2015_aggregate$loss),] 
#colnames(X) <- c("damagecause", "loss")

XX <- X
#XX <- subset(X, loss > 50000000)
YYY1 <- XX

tt3 <- ttheme_default(core=list(fg_params=list(hjust=0, x=0.1)),
                      rowhead=list(fg_params=list(hjust=0, x=0)))

#YYY1<- YYY1[seq(dim(YYY1)[1],1),]
YYY1$loss <- round(YYY1$loss, 0)
#grid.table(YYY1, theme = tt3, rows = NULL)

YYY2 <- YYY1
YYY2$loss <- format(YYY2$loss,big.mark=",",scientific=FALSE)
YYY2$loss <- paste("$", YYY2$loss, sep="")
YYY2$loss <- gsub(" ", "", YYY2$loss, fixed = TRUE)

dd2 <- cbind(YYY2[1:15, ])
colnames(dd2) <- c("year", "loss")
dd2[is.na(dd2)] <- ""

par(mar=c(8, 8.1, 7, 0), family = 'serif', mgp=c(2, 1, 0), las=0)
#levels(YYY1$commodity)[1] <- "ADJ GROSS REVENUE"
barplot(YYY1$loss, names.arg = YYY1$year, las = 3, yaxt="n", col = "lightblue", cex.lab = 1, cex.main = 1, cex.names = 1.5, font.lab = 1, font.axis = 1) 
graphics::title(main = "PNW total insurance loss by year: 2001 to 2022", line = 1, adj = 0, cex.main = 1.5)

pts <- pretty(YYY1$loss / 1000000)
pts2 <- pretty(YYY1$loss)
axis(2, las = 1, cex.axis = 1, at = pts2, labels = paste(pts, "M", sep = ""), cex.axis=1.5)
#abline(v = 14.5, lty = 2, color = "red")

mtext("Insurance Claim Loss ($)", cex = 1.5, side=2, line=5.5)
mtext("Year", cex = 1.5, side=1, line=5)


#htmlTable(dd2,
#          cgroup = c("2001-2015"),
#          n.cgroup = c(2),
#          caption = "PNW total insurance loss by year, 2001-2015",
#          align = "c",
#          rnames = FALSE,
#          css.cell = "padding-left: .5em; padding-right: .5em;")
```

\newpage

```{r message=FALSE, warning=FALSE, error=TRUE, echo = FALSE, fig.width = 8, fig.height = 8, fig.cap=paste("Pacific Northwest agricultural insurance loss by damage cause: 2001 to 2022.")}
#Figure 4

#--barplot of damage causes from 2001 to 2015

palouse_sumloss_1989_2015 <- aggregate(RMA_damage_PNW$loss, list(RMA_damage_PNW$damagecause), FUN = "sum")

palouse_sumloss_1989_2015 <- palouse_sumloss_1989_2015[c(8:nrow(palouse_sumloss_1989_2015)),]

X <- palouse_sumloss_1989_2015
colnames(X) <- c("year", "loss")
#X <- palouse_sumloss_2007_2015_aggregate[order(palouse_sumloss_2007_2015_aggregate$loss),] 
#colnames(X) <- c("damagecause", "loss")

XX <- X
#XX <- subset(X, loss > 50000000)
YYY1 <- palouse_sumloss_1989_2015
colnames(YYY1) <- c("damagecause", "loss")

tt3 <- ttheme_default(core=list(fg_params=list(hjust=0, x=0.1)),
                      rowhead=list(fg_params=list(hjust=0, x=0)))

#YYY1<- YYY1[seq(dim(YYY1)[1],1),]
YYY1$loss <- round(YYY1$loss, 0)
#grid.table(YYY1, theme = tt3, rows = NULL)

YYY1$damagecause <- as.factor(YYY1$damagecause)

levels(YYY1$damagecause)[8] <- "Excessive Moisture"
par(mar=c(14, 8.1, 7, 0), family = 'serif', mgp=c(2, 1, 0), las=0)
#levels(YYY1$commodity)[1] <- "ADJ GROSS REVENUE"

YYY1 <- YYY1[order(-YYY1$loss), ]

YYY1 <- YYY1[1:11,]

YYY1[8,]$loss <- YYY1[8,]$loss + YYY1[11,]$loss
YYY1 <- YYY1[1:10,]

YYY1 <- YYY1[order(YYY1$loss, decreasing = TRUE),]
YYY1$damagecause <- factor(YYY1$damagecause)

levels(YYY1$damagecause)[5] <- "Excessive Moisture"
levels(YYY1$damagecause)[10] <- "Wind"

barplot(YYY1$loss, names.arg = YYY1$damagecause, las = 3, main = "", yaxt="n", col = "lightblue", xlab = "", cex.main = 1, cex.lab = 1, cex.names = 1.5, font.lab = 1, font.axis = 1) 
graphics::title(main = "PNW total insurance loss by damage cause: 2001 to 2022", line = 1, adj = 0, cex.main = 1.5)
pts <- pretty(YYY1$loss / 1000000)
pts2 <- pretty(YYY1$loss)
axis(2, las = 1, cex.axis = 1, at = pts2, labels = paste(pts, "M", sep = ""), cex.axis=1.5)
#abline(v = 14.5, lty = 2, color = "red")

mtext("Damage Cause", cex = 1.5, side=1, line=12.5)
mtext("Insurance Claim Loss ($)", cex = 1.5, side=2, line=5.5)


#htmlTable(dd2,
#          cgroup = c("2001-2015"),
#          n.cgroup = c(2),
#          caption = "PNW total insurance loss by year, 2001-2015",
#          align = "c",
#          rnames = FALSE,
#          css.cell = "padding-left: .5em; padding-right: .5em;")

```

\newpage


```{r message=FALSE, warning=FALSE, error=TRUE, echo = FALSE, fig.width = 8, fig.height = 8, fig.cap=paste("Pacific Northwest agricultural insurance loss per claim: 2001 to 2022.")}

options(scipen = 999)
# PNW_lossperclaim <- ggplot(RMA_damage_PNW, aes(x=factor(year), y=lossperclaim))+
#      geom_boxplot(outlier.shape = NA)+ theme_bw() + scale_y_continuous(limits = c(0, 500000)) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
# 
# IPNW_lossperclaim <- ggplot(RMA_damage_IPNW, aes(x=factor(year), y=lossperclaim))+
#      geom_boxplot(outlier.shape = NA)+ theme_bw() + scale_y_continuous(limits = c(0, 500000))+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

RMA_damage_IPNW_wheat <- subset(RMA_damage_IPNW, commodity == "Wheat" | commodity == "WHEAT")
IPNW_wheat_lossperclaim <- ggplot(RMA_damage_IPNW_wheat, aes(x=factor(year), y=lossperclaim))+
     geom_boxplot(outlier.shape = NA)+ theme_bw() + scale_y_continuous(limits = c(0, 500000))+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

RMA_damage_IPNW_apples <- subset(RMA_damage_IPNW, commodity == "Apples" | commodity == "APPLES")
IPNW_apples_lossperclaim <- ggplot(RMA_damage_IPNW_apples, aes(x=factor(year), y=lossperclaim))+
     geom_boxplot(outlier.shape = NA)+ theme_bw() + scale_y_continuous(limits = c(0, 500000))+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

grid.arrange(IPNW_wheat_lossperclaim,IPNW_apples_lossperclaim, nrow=1, ncol=2)

```


\newpage

```{r message=FALSE, warning=FALSE, error=TRUE, echo = FALSE, fig.width = 8, fig.height = 8, fig.cap=paste("PNW total insurance loss by commodity: 2001 to 2022")}


usePackage("plotly")
usePackage("gridExtra")
usePackage("knitr")
usePackage("kableExtra")
usePackage("htmlTable")
usePackage("stringi")
usePackage("extrafont")


#pie chart for 2007-2015

options(scipen=999)

addNoAnswer <- function(x){
  if(is.factor(x)) return(factor(x, levels=c(levels(x), "No Answer")))
  return(x)
}


palouse_sumloss_1989_2015_lossperacres <- aggregate(RMA_PNW$lossperacre, list( RMA_PNW$year, RMA_PNW$state, RMA_PNW$commodity), FUN = "mean")

colnames(palouse_sumloss_1989_2015_lossperacres) <- c("year", "state", "commodity",  "lossperacres")

palouse_sumloss_1989_2015_lossperacres2 <- aggregate(RMA_PNW$lossperacre, list( RMA_PNW$year, RMA_PNW$damagecause), FUN = "mean")

colnames(palouse_sumloss_1989_2015_lossperacres2) <- c("year", "damagecause", "lossperacres")

palouse_sumloss_1989_2015_lossperacres3 <- aggregate(RMA_PNW$lossperacre, list( RMA_PNW$year, RMA_PNW$damagecause, RMA_PNW$state), FUN = "mean")

colnames(palouse_sumloss_1989_2015_lossperacres3) <- c("year", "damagecause", "state", "lossperacres")

#palouse_sumloss_1989_2015_lossperacres <- subset(palouse_sumloss_1989_2015_lossperacres, lossperacres > 200)
#palouse_sumloss_1989_2015_lossperacres2 <- subset(palouse_sumloss_1989_2015_lossperacres2, lossperacres > 200)

#palouse_sumloss_1989_2015_lossperacres <- subset(palouse_sumloss_1989_2015_lossperacres, commodity != "ADJUSTED GROSS REVENUE")
#palouse_sumloss_1989_2015_lossperacres <- subset(palouse_sumloss_1989_2015_lossperacres, commodity != "Other (Snow/Lightning-Etc.)")

palouse_sumloss_1989_2015_lossperacres2$damagecause <- factor(palouse_sumloss_1989_2015_lossperacres2$damagecause)

palouse_sumloss_1989_2015_lossperacres$commodity <- factor(palouse_sumloss_1989_2015_lossperacres$commodity)


levels(palouse_sumloss_1989_2015_lossperacres$commodity)[14] <- "FORAGE"
levels(palouse_sumloss_1989_2015_lossperacres$commodity)[15] <- "APRICOTS"
levels(palouse_sumloss_1989_2015_lossperacres$commodity)[16] <- "FREESTONE PEACHES"
levels(palouse_sumloss_1989_2015_lossperacres$commodity)[17] <- "NECTARINES"
levels(palouse_sumloss_1989_2015_lossperacres$commodity)[24] <- "PASTURE, RANGELAND"
levels(palouse_sumloss_1989_2015_lossperacres$commodity)[30] <- "BLACKBERRIES"

levels(palouse_sumloss_1989_2015_lossperacres2$damagecause)[8] <- "Excessive Moisture"
levels(palouse_sumloss_1989_2015_lossperacres2$damagecause)[25] <- "Snow/Lightning"
levels(palouse_sumloss_1989_2015_lossperacres2$damagecause)[21] <- "Hurricane"
levels(palouse_sumloss_1989_2015_lossperacres2$damagecause)[24] <- "Mycotoxin"
levels(palouse_sumloss_1989_2015_lossperacres2$damagecause)[22] <- "Lack Irrig Prep"


palouse_sumloss_1989_2015_loss <- aggregate(RMA_PNW$loss, by = list( RMA_PNW$year, RMA_PNW$commodity), FUN = "sum")

colnames(palouse_sumloss_1989_2015_loss) <- c("year", "commodity",  "loss")

palouse_sumloss_1989_2015_loss2 <- aggregate(RMA_PNW$loss, list( RMA_PNW$year, RMA_PNW$damagecause), FUN = "sum")

colnames(palouse_sumloss_1989_2015_loss2) <- c("year", "damagecause", "loss")

palouse_sumloss_1989_2015_loss4 <- aggregate(RMA_PNW$loss, list( RMA_PNW$year, RMA_PNW$damagecause, RMA_PNW$state), FUN = "sum")

colnames(palouse_sumloss_1989_2015_loss4) <- c("year", "damagecause", "state", "loss")





psloss <- palouse_sumloss_1989_2015_loss

palouse_sumloss_1989_2015_loss <- subset(palouse_sumloss_1989_2015_loss, loss > 20000000)
palouse_sumloss_1989_2015_loss2 <- subset(palouse_sumloss_1989_2015_loss2, loss > 20000000)

palouse_sumloss_1989_2015_loss <- subset(palouse_sumloss_1989_2015_loss, commodity != "ADJUSTED GROSS REVENUE")

palouse_sumloss_1989_2015_loss2$damagecause <- factor(palouse_sumloss_1989_2015_loss2$damagecause)

palouse_sumloss_1989_2015_loss$commodity <- factor(palouse_sumloss_1989_2015_loss$commodity)


levels(palouse_sumloss_1989_2015_loss2$damagecause)[4] <- "Excessive Moisture"






palouse_sumloss_1989_2015_count <- aggregate(RMA_PNW$loss, list( RMA_PNW$year, RMA_PNW$state, RMA_PNW$commodity), FUN = "length")

colnames(palouse_sumloss_1989_2015_count) <- c("year", "state", "commodity",  "loss")

palouse_sumloss_1989_2015_count2<- aggregate(RMA_PNW$loss, list( RMA_PNW$year, RMA_PNW$damagecause, RMA_PNW$state), FUN = "length")

colnames(palouse_sumloss_1989_2015_count2) <- c("year", "damagecause", "state", "loss")


palouse_sumloss_1989_2015_loss3<- aggregate(RMA_PNW$loss, list( RMA_PNW$year, RMA_PNW$commodity, RMA_PNW$state), FUN = "sum")

colnames(palouse_sumloss_1989_2015_loss3) <- c("year", "commodity", "state", "loss")

#-boxplots of count data for all commodities and all damage causes- 1989-2015

palouse_sumloss_1989_2015_count <- subset(palouse_sumloss_1989_2015_count, loss > 50)
palouse_sumloss_1989_2015_count2 <- subset(palouse_sumloss_1989_2015_count2, loss > 50)

palouse_sumloss_1989_2015_count <- subset(palouse_sumloss_1989_2015_count, commodity != "ADJUSTED GROSS REVENUE")

palouse_sumloss_1989_2015_count2$damagecause <- factor(palouse_sumloss_1989_2015_count2$damagecause)

palouse_sumloss_1989_2015_count$commodity <- factor(palouse_sumloss_1989_2015_count$commodity)


levels(palouse_sumloss_1989_2015_count2$damagecause)[5] <- "Excessive Moisture"
levels(palouse_sumloss_1989_2015_count2$damagecause)[11] <- "Other"

```



\newpage
```{r message=FALSE, warning=FALSE, error=TRUE, echo = FALSE, fig.width = 8, fig.height = 8, fig.cap=paste("Pacific Northwest agricultural insurance loss by commodity: 2001 to 2022.")}
#Figure 5

RMA_PNW_com <- RMA_PNW
RMA_PNW_com$commodity <- tolower(RMA_PNW_com$commodity)
RMA_PNW_com$commodity <- stri_trans_general(RMA_PNW_com$commodity, id = "Title")

palouse_sumloss_1989_2015 <- aggregate(RMA_PNW_com$loss, list(RMA_PNW_com$commodity), FUN = "sum")

X <- palouse_sumloss_1989_2015
colnames(X) <- c("commodity", "loss")
#X <- palouse_sumloss_2007_2015_aggregate[order(palouse_sumloss_2007_2015_aggregate$loss),] 
#colnames(X) <- c("damagecause", "loss")

XX <- X
XX <- subset(X, loss > 50000000)
YYY1 <- XX

tt3 <- ttheme_default(core=list(fg_params=list(hjust=0, x=0.1)),
                      rowhead=list(fg_params=list(hjust=0, x=0)))

#YYY1<- YYY1[seq(dim(YYY1)[1],1),]
YYY1$loss <- round(YYY1$loss, 0)
#grid.table(YYY1, theme = tt3, rows = NULL)

YYY2 <- YYY1
YYY2$loss <- format(YYY2$loss,big.mark=",",scientific=FALSE)
YYY2$loss <- paste("$", YYY2$loss, sep="")
YYY2$loss <- gsub(" ", "", YYY2$loss, fixed = TRUE)

dd2 <- cbind(YYY2[1:14, ])
colnames(dd2) <- c("commodity", "loss")
dd2[is.na(dd2)] <- ""

YYY1$commodity <- tolower(YYY1$commodity)
YYY1$commodity <- stri_trans_general(YYY1$commodity, id = "Title")
YYY1$commodity <- factor(YYY1$commodity)


par(mar=c(10, 8.1, 6, 0), family = 'serif', mgp=c(2, 1, 0), las=0)
levels(YYY1$commodity)[1] <- "Gross Revenue"
levels(YYY1$commodity)[2] <- "Other Crops"


YYY1 <- YYY1[order(-YYY1$loss), ]
YYY1 <- YYY1[-c(2:3),]
YYY1$commodity <- factor(YYY1$commodity)


barplot(YYY1$loss, names.arg = YYY1$commodity, las = 3, yaxt="n", col = "lightblue", cex.main = 1.5, cex.lab = 1, cex.names = 1.5, font.lab = 1, font.axis = 1) 
graphics::title(main = "PNW total insurance loss by commodity: 2001 to 2022", line = 1, adj = 0, cex.main = 1.5)
pts <- pretty(YYY1$loss / 1000000)
pts2 <- pretty(YYY1$loss)
axis(2, las = 1, cex.axis = 1, at = pts2, labels = paste(pts, "M", sep = ""), cex.axis=1.5)
#abline(v = 14.5, lty = 2, color = "red")

mtext("Insurance Claim Loss ($)", cex = 1.5, side=2, line=5.5)
mtext("Commodity", cex = 1.5, side=1, line=8)


#htmlTable(dd2,
#          cgroup = c("2001-2015"),
#          n.cgroup = c(2),
#          caption = "PNW total insurance loss by year, 2001-2015",
#          align = "c",
#          rnames = FALSE,
#          css.cell = "padding-left: .5em; padding-right: .5em;")


```


\newpage

```{r message=FALSE, warning=FALSE, error=TRUE, echo = FALSE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: Biplot of principal components for insurance loss for the entire PNW, by county, with commodity as the factor loadings. Bottom panel: Scree plot. Data from 2001 to 2022 is used.")}

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2")
usePackage("extrafont")

#PNW by commodity

RMA_damage_PNW_com <- RMA_damage_PNW[RMA_damage_PNW$commodity == c("Wheat", "Apples", "Cherries", "Potatoes", "Dry Peas", "Grapes", "Barley", "Sugar Beets"),]

RMA_damage_PNW_transformed_commodity <- dcast(RMA_damage_PNW_com, county + state ~ commodity, value.var = c("loss"), sum)
RMA_damage_PNW_transformed_commodity$statecounty <- paste(RMA_damage_PNW_transformed_commodity$county, "_", RMA_damage_PNW_transformed_commodity$state, sep="")
rownames(RMA_damage_PNW_transformed_commodity) <- RMA_damage_PNW_transformed_commodity$statecounty

RMA_damage_PNW_exogenous_commodity <- RMA_damage_PNW_transformed_commodity[4:ncol(RMA_damage_PNW_transformed_commodity)-1]
#RMA_damage_PNW_exogenous_commodity <- RMA_damage_PNW_transformed_commodity[,3:41]

#RMA_damage_PNW_exogenous_commodity <- RMA_damage_PNW_exogenous_commodity[-1]
#RMA_damage_PNW_exogenous_commodity <- RMA_damage_PNW_exogenous_commodity[-20]
#RMA_damage_PNW_exogenous_commodity <- RMA_damage_PNW_exogenous_commodity[-37]


fit <- princomp(RMA_damage_PNW_exogenous_commodity, cor=TRUE)
#ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_commodity, scale = TRUE, center = TRUE), data = RMA_damage_PNW_transformed_commodity, colour = 'county', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 3, legend = FALSE) + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_commodity)) + theme(legend.position = "none")


plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_commodity, center = TRUE, scale = TRUE),   data = RMA_damage_PNW_transformed_commodity, loadings = TRUE, loadings.label = TRUE, colour = 'state', loadings.label.size  = 3, label = FALSE, label.size = 2, legend = TRUE, loadings.label.repel=TRUE)  + theme_bw()   +  theme(legend.position = "none") + ggtitle("Principal Components Analysis: PNW insurance loss by county with commodity loadings ") + theme(plot.title = element_text(size =12)) + theme(axis.title.y = element_text(size=12), axis.title.x = element_text(size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1), axis.text.y = element_text(size=rel(1.5), hjust = 1))

#colour = 'state',

#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)
prop_varex$comp_factor <- as.factor(prop_varex$comp)


p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1)) + theme(text=element_text(family= "Montserrat"))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5), text=element_text(family= "Montserrat")) +
xlab("comp") + theme_bw() + stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot: PNW insurance loss by county with commodity loadings", x = "principal components", y = "% explained variance") 

#+ panel.border = element_rect(colour = "black", fill=NA, size=1)


#grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))


```
\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of principal components of insurance loss, for the entire PNW, for all commodities by year, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 to 2022 is used.")}

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2")
usePackage("grid")
usePackage("gridExtra")

RMA_damage_PNW_com <- RMA_damage_PNW[RMA_damage_PNW$damagecause == c("Drought", "Decline in Price", "Heat", "Freeze", "Frost", "Cold Wet Weather", "Hail", "Excess Moisture/Precipitation/Rain", "Cold Winter", "Wind/Excess Wind"),]

RMA_damage_PNW_transformed_year <- dcast(RMA_damage_PNW_com, year ~ damagecause, value.var = c("loss"), sum)
#RMA_damage_PNW_transformed_year$statecounty <- paste(RMA_damage_PNW_transformed_year$county, "_", RMA_damage_PNW_transformed_year$state, sep="")
rownames(RMA_damage_PNW_transformed_year) <- RMA_damage_PNW_transformed_year$year
RMA_damage_PNW_exogenous_year_factor <- RMA_damage_PNW_transformed_year[,2:11]

#PNW using year as group - with damage cause as loadings
#fit <- princomp(RMA_damage_PNW_exogenous_year, cor=TRUE, scale = TRUE, center = TRUE)

#RMA_damage_PNW_exogenous_year_factor <- cbind(RMA_damage_PNW_exogenous_year[,2:3], RMA_damage_PNW_exogenous_year[6], RMA_damage_PNW_exogenous_year[8], RMA_damage_PNW_exogenous_year[,11:14], RMA_damage_PNW_exogenous_year[16:17])

fit <- princomp(RMA_damage_PNW_exogenous_year_factor, cor=TRUE)

plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_year_factor, center = TRUE, scale = TRUE),  data = RMA_damage_PNW_exogenous_year_factor, loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 2, legend = FALSE, loadings.label.hjust = -.1, loadings.label.repel=T)  + theme_bw()   + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_year_factor)) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: PNW insurance loss by year with damage cause loadings ") + theme(plot.title = element_text(size =12)) + theme(axis.title.y = element_text( size=12), axis.title.x = element_text(size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1), axis.text.y = element_text(size=rel(1.5), hjust = 1)) 

#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot: PNW insurance loss by year with damage cause loadings", x = "principal components", y = "% explained variance")

#grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))

```
\newpage
<br></br>


```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of principal components for insurance loss for the entire PNW, for all commodities by month, with damage cause as the factor loadings.  Bottom panel: Scree plot Data from 2001 is 2022 is used.")}

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2")

#PNW damage cause as loadings using months as groups

#remove negatives related to lossperacre calc
RMA_PNW_noneg <- subset(RMA_PNW, lossperacre != Inf)
RMA_PNW_noneg2 <- subset(RMA_PNW_noneg, lossperacre >=0 )
RMA_PNW_noneg3 <- subset(RMA_PNW_noneg2, month !="" )

RMA_PNW_noneg3_com <- RMA_PNW_noneg3[RMA_PNW_noneg3$damagecause == c("Drought", "Decline in Price", "Heat", "Freeze", "Frost", "Cold Wet Weather", "Hail", "Excess Moisture/Precipitation/Rain", "Cold Winter", "Plant Disease"),]

RMA_damage_PNW_transformed_month <- dcast(RMA_PNW_noneg3_com, month ~ damagecause, value.var = c("loss"), mean)

is.nan.data.frame <- function(x)
do.call(cbind, lapply(x, is.nan))

RMA_damage_PNW_transformed_month[is.nan(RMA_damage_PNW_transformed_month)] <- 0

#RMA_damage_PNW_transformed_year$statecounty <- paste(RMA_damage_PNW_transformed_year$county, "_", RMA_damage_PNW_transformed_year$state, sep="")
rownames(RMA_damage_PNW_transformed_month) <- RMA_damage_PNW_transformed_month$month
RMA_damage_PNW_exogenous_month <- RMA_damage_PNW_transformed_month[,2:10]

#RMA_damage_PNW_exogenous_month <- cbind(RMA_damage_PNW_exogenous_month[,2:3], RMA_damage_PNW_exogenous_month[6], RMA_damage_PNW_exogenous_month[8], RMA_damage_PNW_exogenous_month[,11:14], RMA_damage_PNW_exogenous_month[16:17])


#ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_year, center = TRUE, scale = TRUE), data = RMA_damage_PNW_transformed_year, colour = 'year', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 3, legend = FALSE) + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_year)) + theme(legend.position = "none")

#PNW using month as group - with damage cause as loadings
fit <- princomp(RMA_damage_PNW_exogenous_month, cor=TRUE)
plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_month, center = TRUE, scale = TRUE),  data = RMA_damage_PNW_transformed_month,  loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 2, legend = FALSE)  + theme_bw()   + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_month)) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: PNW insurance loss by month with damage cause loadings ") + theme(plot.title = element_text(size =12)) + theme(axis.title.y = element_text(size=12), axis.title.x = element_text(size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif"))
#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot: PNW insurance loss, for all commodities by month, with damage cause \n as the factor loadings. ", x = "principal components", y = "% explained variance")

#grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))

```
\newpage
<br></br>


```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of insurance loss for the entire PNW, for wheat by year, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used.")}

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2")

#PNW damage cause as loadings using years as groups, JUST FOR WHEAT



RMA_damage_PNW_com$commodity <- tolower(RMA_damage_PNW_com$commodity)
RMA_damage_PNW_com$commodity <- stri_trans_general(RMA_damage_PNW_com$commodity, id = "Title")
RMA_damage_PNW_com$commodity <- factor(RMA_damage_PNW_com$commodity)


RMA_damage_PNW_wheat_year <- subset(RMA_damage_PNW_com, commodity == "Wheat")
RMA_damage_PNW_transformed_wheat_year <- dcast(RMA_damage_PNW_wheat_year, year ~ damagecause, value.var = c("loss"), sum)
#RMA_damage_PNW_transformed_year$statecounty <- paste(RMA_damage_PNW_transformed_year$county, "_", RMA_damage_PNW_transformed_year$state, sep="")
rownames(RMA_damage_PNW_transformed_wheat_year) <- RMA_damage_PNW_transformed_wheat_year$year
RMA_damage_PNW_exogenous_wheat_year <- RMA_damage_PNW_transformed_wheat_year
par(mar=c(8, 6.1, 4, 2.1), family = 'serif', mgp=c(4, 1, 0), las=0)

#PNW using year as group - with damage cause as loadings: JUST FOR WHEAT

#RMA_damage_PNW_exogenous_wheat_year_factor <- cbind(RMA_damage_PNW_exogenous_wheat_year[,3:4], RMA_damage_PNW_exogenous_wheat_year[7], RMA_damage_PNW_exogenous_wheat_year[9], RMA_damage_PNW_exogenous_wheat_year[,12:15], RMA_damage_PNW_exogenous_wheat_year[17:18])

RMA_damage_PNW_exogenous_wheat_year <- RMA_damage_PNW_exogenous_wheat_year[2:11]
fit <- princomp(RMA_damage_PNW_exogenous_wheat_year, cor=TRUE)

plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_wheat_year, center = TRUE, scale = TRUE), data = RMA_damage_PNW_exogenous_wheat_year, loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 2, legend = FALSE)  + theme_bw()   + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_wheat_year)) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: PNW wheat insurance loss by year with \ndamage cause loadings ") + theme(plot.title = element_text(size =12, family = "Serif")) + theme(axis.title.y = element_text(family = "Serif", size=12), axis.title.x = element_text(family = "Serif", size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1, family = "Serif"), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif"))

#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot: insurance loss for the entire PNW, for wheat by year, with \ndamage cause as the factor loadings.", x = "principal components", y = "% explained variance")

#grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))


```
\newpage
<br></br>


```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Inland Pacific Northwest agricultural insurance loss by year, 2001 to 2022.")}

palouse_sumloss_aggregate <- subset(RMA_damage_IPNW, year >= 2001 & year <= 2022) 

#palouse_sumloss_1989_2015 <- subset(palouse_sumloss_aggregate, commodity == "WHEAT")

palouse_sumloss_1989_2015 <- aggregate(RMA_damage_IPNW$loss, list(RMA_damage_IPNW$year), FUN = "sum")

colnames(palouse_sumloss_1989_2015) <- c("year", "loss")

#palouse_sumloss_2007_2015_aggregate <- palouse_sumloss_1989_2015[order(palouse_sumloss_1989_2015$loss),] 

X <- palouse_sumloss_1989_2015

#X <- palouse_sumloss_2007_2015_aggregate[order(palouse_sumloss_2007_2015_aggregate$loss),] 
#colnames(X) <- c("damagecause", "loss")

XX <- X
#XX <- subset(X, loss > 1000000)
YYY1 <- XX

tt3 <- ttheme_default(core=list(fg_params=list(hjust=0, x=0.1)),
                      rowhead=list(fg_params=list(hjust=0, x=0)))

#YYY1<- YYY1[seq(dim(YYY1)[1],1),]
YYY1$loss <- round(YYY1$loss, 0)
#grid.table(YYY1, theme = tt3, rows = NULL)

YYY2 <- YYY1
YYY2$loss <- format(YYY2$loss,big.mark=",",scientific=FALSE)
YYY2$loss <- paste("$", YYY2$loss, sep="")
YYY2$loss <- gsub(" ", "", YYY2$loss, fixed = TRUE)

dd2 <- cbind(YYY2[1:15, ])
colnames(dd2) <- c("year", "loss")
dd2[is.na(dd2)] <- ""


par(mar=c(9, 8.1, 5, 2.1), family = 'serif', mgp=c(6, 1, 0), las=0)
#levels(YYY1$commodity)[1] <- "ADJ GROSS REVENUE"
barplot(YYY1$loss, names.arg = YYY1$year, las = 3, cex.names = 1.5, yaxt="n", col = "lightblue") 
pts <- pretty(YYY1$loss / 1000000)
pts2 <- pretty(YYY1$loss)
axis(2, las = 1.5, at = pts2, labels = paste(pts, "M", sep = ""), cex.axis=1.5)
graphics::title(main="IPNW total insurance loss: 2001 to 2022", line = 1, cex.main = 1.5, adj = 0)
graphics::title(xlab="Years", line=6, cex.lab = 1.5)
graphics::title(ylab="Insurance claim loss ($)", line=5.5, cex.lab = 1.5)

#abline(v = 14.5, lty = 2, color = "red")

```

\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Inland Pacific Northwest agricultural insurance loss by damage cause, 2001 to 2022.")}

#pie chart for 2007-2015

options(scipen=999)

addNoAnswer <- function(x){
  if(is.factor(x)) return(factor(x, levels=c(levels(x), "No Answer")))
  return(x)
}

#palouse_sumloss_1989_2015 <- subset(palouse_sumloss_aggregate, commodity == "WHEAT")
palouse_sumloss_1989_2015 <- aggregate(palouse_sumloss_aggregate$loss, list(palouse_sumloss_aggregate$damagecause), FUN = "sum")
colnames(palouse_sumloss_1989_2015) <- c("commodity", "loss")

palouse_sumloss_2007_2015_aggregate <- palouse_sumloss_1989_2015[order(palouse_sumloss_1989_2015$loss),] 
X <- palouse_sumloss_2007_2015_aggregate[order(palouse_sumloss_2007_2015_aggregate$loss),] 
#colnames(X) <- c("damagecause", "loss")
XX <- subset(X, loss > 10000000)
YYY1 <- XX

tt3 <- ttheme_default(core=list(fg_params=list(hjust=0, x=0.1)),
                      rowhead=list(fg_params=list(hjust=0, x=0)))


YYY1<- YYY1[seq(dim(YYY1)[1],1),]

YYY2 <- YYY1

YYY2$commodity <- as.factor(YYY2$commodity)


YYY2$loss[7] <- YYY2$loss[7]+YYY2$loss[13]

YYY2 <- YYY2[c(1:9,11, 16:22),]
YYY2$commodity <- factor(YYY2$commodity)

levels(YYY2$commodity)[5] <- "Excessive Moisture"
levels(YYY2$commodity)[14] <- "Snow/Lightning"
YYY2$loss <- format(YYY2$loss,big.mark=",",scientific=FALSE)
YYY2$loss <- paste("$", YYY2$loss, sep="")
YYY2$loss <- gsub(" ", "", YYY2$loss, fixed = TRUE)
colnames(YYY2) <- c("damage cause", "loss")

#YYY2 <- subset(YYY2, commodity != "ADJUSTED GROSS REVENUE")


#grid.table(YYY1, theme = tt3)
#kable(YYY1, format = "html", row.names = FALSE) %>% kable_styling(bootstrap_options = c("striped", "hover", "bordered"), full_width = FALSE)


YYY1$commodity <- as.factor(YYY1$commodity)

YYY1$loss[7] <- YYY1$loss[7]+YYY1$loss[13]

YYY1 <- YYY1[c(1:9,11, 16:22),]
YYY1$commodity <- factor(YYY1$commodity)
par(mar=c(10, 9.8, 8, 2.1), family = 'serif', mgp=c(2, 1, 0), las=0)

levels(YYY1$commodity)[5] <- "Moisture"
levels(YYY1$commodity)[14] <- "Snow/Lightning"
levels(YYY1$commodity)[17] <- "Wind"
levels(YYY1$commodity)[1] <- "Cold Weather"

YYY1 <- YYY1[1:10,]

#levels(YYY1$commodity)[10] <- "Excessive Moisture"
barplot(YYY1$loss, names.arg = YYY1$commodity, cex.names = 1.5, las = 3, yaxt="n", col = "lightblue")
pts <- pretty(YYY1$loss / 1000000)
pts2 <- pretty(YYY1$loss)
axis(2, las = 1, at = pts2, cex.axis = 1, labels = paste(pts, "M", sep = ""), cex.axis = 1.5)

#mtext(text = "Damage Cause",
#      side = 1,#side 1 = bottom
#      line = 7.8)

graphics::title(main="IPNW insurance loss by damage cause: 2001 to 2022", line = 1, cex.main = 1.5, adj = 0)
graphics::title(xlab="Damage Cause", line=8.5, cex.lab = 1.5)
graphics::title(ylab="Insurance claim loss ($)", line=5.5, cex.lab = 1.5)

#htmlTable(YYY2,
#          cgroup = c("2001-2015"),
#          caption = "IPNW region total insurance loss by damage cause, 2001-2015",
#          align = "c",
#          rnames = FALSE,
#          css.cell = "padding-left: .5em; padding-right: .5em;")


```

\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Inland Pacific Northwest agricultural insurance loss by commodity, 2001 to 2022.")}

#htmlTable(dd2,
#          cgroup = c("2001-2015"),
#          n.cgroup = c(2),
 #         caption = "IPNW region total insurance loss by year, 2001-2015",
#          align = "c",
 #         rnames = FALSE,
#          css.cell = "padding-left: .5em; padding-right: .5em;")





usePackage("plotly")
usePackage("gridExtra")
usePackage("stringi")

#pie chart for 2007-2015

options(scipen=999)

addNoAnswer <- function(x){
  if(is.factor(x)) return(factor(x, levels=c(levels(x), "No Answer")))
  return(x)
}

palouse_sumloss_aggregate_com <- palouse_sumloss_aggregate
palouse_sumloss_aggregate_com$commodity <- tolower(palouse_sumloss_aggregate_com$commodity)
palouse_sumloss_aggregate_com$commodity <- stri_trans_general(palouse_sumloss_aggregate_com$commodity, id = "Title")
palouse_sumloss_aggregate_com$commodity <- factor(palouse_sumloss_aggregate_com$commodity)

#palouse_sumloss_aggregate <- subset(palouse_sumloss_aggregate, year >= 1989 & year <= 201)
palouse_sumloss_1989_2015 <- aggregate(palouse_sumloss_aggregate_com$loss, list(palouse_sumloss_aggregate_com$commodity), FUN = "sum")
colnames(palouse_sumloss_1989_2015) <- c("commodity", "loss")

palouse_sumloss_2007_2015_aggregate <- palouse_sumloss_1989_2015[order(palouse_sumloss_1989_2015$loss),] 
X <- palouse_sumloss_2007_2015_aggregate[order(palouse_sumloss_2007_2015_aggregate$loss),] 
#colnames(X) <- c("damagecause", "loss")
#XX <- subset(X, loss > 15000000)
YYY1 <- X

tt3 <- ttheme_default(core=list(fg_params=list(hjust=0, x=0.1)),
                      rowhead=list(fg_params=list(hjust=0, x=0)))

YYY1<- YYY1[seq(dim(YYY1)[1],1),]
#grid.table(YYY1, theme = tt3)


YYY2 <- YYY1
YYY2$loss <- format(YYY2$loss,big.mark=",",scientific=FALSE)
YYY2$loss <- paste("$", YYY2$loss, sep="")
YYY2$loss <- gsub(" ", "", YYY2$loss, fixed = TRUE)

YYY2 <- subset(YYY2, commodity != "Adjusted Gross Revenue")
YYY2 <- subset(YYY2, commodity != "Whole Farm Revenue Protection")
levels(YYY2$commodity)[levels(YYY2$commodity)=="Pasture,Rangeland,Forage"] <- "Forage"


#kable(YYY1, format = "html", row.names = FALSE) %>% kable_styling(bootstrap_options = c("striped", "hover", "bordered"), full_width = FALSE)


YYY1 <- subset(YYY1, commodity != "Adjusted Gross Revenue")
#YYY1 <- subset(YYY1, commodity != "ADJUSTED GROSS REVENUE-LITE")
YYY1 <- subset(YYY1, commodity != "Whole Farm Revenue Protection")
levels(YYY1$commodity)[levels(YYY1$commodity)=="Pasture,Rangeland,Forage"] <- "Forage"

par(mar=c(8, 9.1, 6, 2.1), family = 'serif', mgp=c(6, 1, 0), las=0)

YYY1$commodity<- tolower(YYY1$commodity)
YYY1$commodity <- stri_trans_general(YYY1$commodity, id = "Title")

YYY1 <- YYY1[1:8,]

#levels(YYY1$commodity)[1] <- "ADJ GROSS REVENUE"
barplot(YYY1$loss, names.arg = YYY1$commodity, cex.names = 1.5, las = 3, yaxt="n", cex.lab = 1, col = "lightblue") 
pts <- pretty(YYY1$loss / 1000000)
pts2 <- pretty(YYY1$loss)
axis(2, las = 1, at = pts2, cex.axis = 1, labels = paste(pts, "M", sep = ""), cex.axis=1.5)

graphics::title(main="IPNW total insurance loss by commodity: 2001 to 2022", line = 1, cex.main = 1.5, adj = 0)
graphics::title(xlab="Commodity", line=6.5, cex.lab = 1.5)
graphics::title(ylab="Insurance claim loss ($)", line=5.5, cex.lab = 1.5)


#htmlTable(YYY2,
#          cgroup = c("2001-2015"),
#          caption = "IPNW region total insurance loss by commodity, 2001-2015",
#          align = "c",
#          rnames = FALSE,
#          css.cell = "padding-left: .5em; padding-right: .5em;")

```

\newpage

```{r message=FALSE, warning=FALSE, echo = FALSE, error=TRUE}

#--reduce years to 2001-2015

RMA_damage_IPNW_com <- RMA_damage_IPNW
RMA_damage_IPNW_com$commodity <- tolower(RMA_damage_IPNW_com$commodity)
RMA_damage_IPNW_com$commodity <- stri_trans_general(RMA_damage_IPNW_com$commodity, id = "Title")
RMA_damage_IPNW_com$commodity <- factor(RMA_damage_IPNW_com$commodity)


xxyear <- subset(RMA_damage_IPNW_com, year >= 2001)
xxyear$log10_loss <- log10(xxyear$loss)

#missing data review of data from 2001-2015 for all commodities and all damage causes
#ezDesign(xxyear, year, damagecause)
#ezDesign(xxyear, county, damagecause)
#ezDesign(xxyear, county, commodity)

#Now lets subset by the four main commodities of interest.
#--subset to four commodities - Barley, Wheat, Apples, and Dry Peas
xx <- subset(xxyear, commodity == "Barley" | commodity == "Wheat" | 
               commodity == "Apples" | commodity == "Dry Peas" | commodity == "Cherries")

#--subset to a select set of damage causes - Drought, Heat, Hail, Frost, Freeze, Excessive Moisture, Cold Winter, Cold Weather, and Decline in Price
xxx <- subset(xx, damagecause == "Drought" | damagecause == "Heat" | 
                damagecause == "Hail" | damagecause == "Frost" | damagecause == "Freeze" | damagecause == "Excessive Moisture/Precip/Rain" | damagecause == "Cold Winter" | 
                damagecause == "Cold Wet Weather" | damagecause == "Decline in Price")


#examine missing data after narrowing commodities and damage causes to the most relevant
#ezDesign(xxx, year, damagecause)
#ezDesign(xxx, county, year)
#ezDesign(xxx, county, damagecause)
#ezDesign(xxx, year, commodity)
#ezDesign(xxx, county, commodity)

#palouse_sumloss_aggregate <- xxx

#now lets divide the data into four different model files that are commodity specific.
xxx_wheat <- subset(xxx, commodity == "Wheat")
xxx_apples <- subset(xxx, commodity == "Apples")
xxx_cherries <- subset(xxx, commodity == "Cherries")
xxx_drypeas <- subset(xxx, commodity == "Dry Peas")
xxx_barley <- subset(xxx, commodity == "Barley")

palouse_sumloss_aggregate_apples <- xxx_apples
palouse_sumloss_aggregate_cherries <- xxx_cherries
palouse_sumloss_aggregate_drypeas <- xxx_drypeas
palouse_sumloss_aggregate_wheat <- xxx_wheat
palouse_sumloss_aggregate_barley <- xxx_barley

#-remove counties for each commodity file that have considerable missing data.  
#the reasoning is that those counties that have very little of the commodity in question
#may be inappropriate to fill in data as zero.  While there are sporadic commodity loss
#for these counties, there is a high liklihood that this commodity may NOT be grown
#for the missing years, vs just no commodity loss claims.  So we remove these select
#counties given the EzDesign data review of each commodity file

palouse_sumloss_aggregate_apples <- subset(palouse_sumloss_aggregate_apples, 
                                           county != "Columbia" & county != "Wasco")
palouse_sumloss_aggregate_apples$county <- 
  factor(palouse_sumloss_aggregate_apples$county)

palouse_sumloss_aggregate_apples <- 
  subset(palouse_sumloss_aggregate_apples, damagecause != "Drought")
palouse_sumloss_aggregate_apples$damagecause <- 
  factor(palouse_sumloss_aggregate_apples$damagecause)


palouse_sumloss_aggregate_wheat <- 
  subset(palouse_sumloss_aggregate_wheat, county != "Kootenai")
palouse_sumloss_aggregate_wheat$county <- 
  factor(palouse_sumloss_aggregate_wheat$county)

palouse_sumloss_aggregate_cherries <- 
  subset(palouse_sumloss_aggregate_cherries, county != "Adams" & county != "Union")
palouse_sumloss_aggregate_cherries$county <- 
  factor(palouse_sumloss_aggregate_cherries$county)

palouse_sumloss_aggregate_drypeas <- 
  subset(palouse_sumloss_aggregate_drypeas, county != "Douglas" 
         & county != "Gilliam" & county != "Adams")
palouse_sumloss_aggregate_drypeas$county <- 
  factor(palouse_sumloss_aggregate_drypeas$county)

palouse_sumloss_aggregate_barley <- 
  subset(palouse_sumloss_aggregate_barley, county != "Douglas" & county != "Garfield"
         & county != "Gilliam" & county != "Adams" & county != "Kootenai" & county != "Benton")
palouse_sumloss_aggregate_barley$county <- 
  factor(palouse_sumloss_aggregate_barley$county)

palouse_sumloss_aggregate_barley <- 
  subset(palouse_sumloss_aggregate_barley, damagecause != "Cold Winter")
palouse_sumloss_aggregate_barley$damagecause <- 
  factor(palouse_sumloss_aggregate_barley$damagecause)

#re-factor 

palouse_sumloss_aggregate_apples$damagecause <- factor(palouse_sumloss_aggregate_apples$damagecause)
palouse_sumloss_aggregate_wheat$damagecause <- factor(palouse_sumloss_aggregate_wheat$damagecause)
palouse_sumloss_aggregate_cherries$damagecause <- factor(palouse_sumloss_aggregate_cherries$damagecause)
palouse_sumloss_aggregate_drypeas$damagecause <- factor(palouse_sumloss_aggregate_drypeas$damagecause)
palouse_sumloss_aggregate_barley$damagecause <- factor(palouse_sumloss_aggregate_barley$damagecause)

```

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Inland Pacific Northwest agricultural insurance loss for apples, by county, 2001 to 2022.")}

usePackage("plotly")
usePackage("gridExtra")

#pie chart for 2007-2015

options(scipen=999)

addNoAnswer <- function(x){
  if(is.factor(x)) return(factor(x, levels=c(levels(x), "No Answer")))
  return(x)
}

APPLES <- subset(RMA_damage_IPNW_com, commodity == "Apples")

palouse_sumloss_1989_2015 <- aggregate(APPLES$loss, list(APPLES$damagecause), FUN = "sum")
colnames(palouse_sumloss_1989_2015) <- c("damagecause", "loss")

palouse_sumloss_2007_2015_aggregate <- palouse_sumloss_1989_2015[order(palouse_sumloss_1989_2015$loss),] 
X <- palouse_sumloss_2007_2015_aggregate[order(palouse_sumloss_2007_2015_aggregate$loss),] 
#colnames(X) <- c("damagecause", "loss")
summary_dataset <- X %>% filter(loss < 1500000) 
summary_dataset2 <- X %>% filter(loss > 1500000) 
colnames(summary_dataset) <- c("damagecause", "loss")
Y <- sum(summary_dataset$loss)
YY <- c("Other", Y)
YY <- data.frame(t(YY))
colnames(YY) <- c("damagecause", "loss")
YYY <- rbind(summary_dataset2, YY)
#factor(YYY)

YYY <- as.data.frame(lapply(YYY, addNoAnswer))
YYY$loss <- as.numeric(as.character(YYY$loss))
YYY1<- YYY[order(YYY$loss),] 
YYY1 <- YYY1[ nrow(YYY1):1, ]

#--year
palouse_sumloss_year <- aggregate(APPLES$loss, list(APPLES$year), FUN = "sum")
colnames(palouse_sumloss_year) <- c("year", "loss")

#--county

palouse_sumloss_county <- aggregate(APPLES$loss, list(APPLES$county), FUN = "sum")
colnames(palouse_sumloss_county) <- c("county", "loss")

p <- plot_ly(YYY, labels = ~damagecause, values = ~loss, type = 'pie',
             textposition = 'inside',
             textinfo = 'label+percent',
             insidetextfont = list(color = '#FFFFFF'),
             hoverinfo = 'text',
             text = ~paste('$', loss, ' million'),
             marker = list(colors = colors,
                           line = list(color = '#FFFFFF', width = 1)),
             #The 'pull' attribute can also be used to create space between the sectors
             showlegend = TRUE) %>%
  layout(title = 'Top Damage Causes for APPLES, Palouse Region: 1989-2015',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

p <- plot_ly(palouse_sumloss_year, labels = ~year, values = ~loss, type = 'pie',
             textposition = 'inside',
             textinfo = 'label+percent',
             insidetextfont = list(color = '#FFFFFF'),
             hoverinfo = 'text',
             text = ~paste('$', loss, ' million'),
             marker = list(colors = colors,
                           line = list(color = '#FFFFFF', width = 1)),
             #The 'pull' attribute can also be used to create space between the sectors
             showlegend = TRUE) %>%
  layout(title = 'Top Years for APPLES, Palouse Region: 1989-2015',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

p <- plot_ly(palouse_sumloss_county, labels = ~county, values = ~loss, type = 'pie',
             textposition = 'inside',
             textinfo = 'label+percent',
             insidetextfont = list(color = '#FFFFFF'),
             hoverinfo = 'text',
             text = ~paste('$', loss, ' million'),
             marker = list(colors = colors,
                           line = list(color = '#FFFFFF', width = 1)),
             #The 'pull' attribute can also be used to create space between the sectors
             showlegend = TRUE) %>%
  layout(title = 'Top Counties for APPLES, Palouse Region: 1989-2015',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

           
#p

YYY2 <- YYY1
YYY2$loss <- round(YYY2$loss, digits = 0)
YYY2$loss <- format(YYY2$loss,big.mark=",",scientific=FALSE)
YYY2$loss <- paste("$", YYY2$loss, sep="")
YYY2$loss <- gsub(" ", "", YYY2$loss, fixed = TRUE)

#--year
palouse_sumloss_year <- aggregate(APPLES$loss, list(APPLES$year), FUN = "sum")
colnames(palouse_sumloss_year) <- c("year", "loss")

#--county

palouse_sumloss_county <- aggregate(APPLES$loss, list(APPLES$county), FUN = "sum")
colnames(palouse_sumloss_county) <- c("county", "loss")

# par(mar=c(12.5, 6.1, 4, 2.1), family = 'serif', mgp=c(4, 1, 0), las=0)
# barplot(palouse_sumloss_county$loss, names.arg = palouse_sumloss_county$county, cex.names = 1, las = 3, yaxt="n",  col = "lightblue")
# pts <- pretty(palouse_sumloss_county$loss / 1000000)
# pts2 <- pretty(palouse_sumloss_county$loss)
# axis(2, las = 1, at = pts2, cex.axis = 1, labels = paste(pts, "M", sep = ""))
# 
# #mtext(text = "Damage Cause",
# #      side = 1,#side 1 = bottom
# #      line = 8)
# 
# graphics::title(main="Apples insurance loss by commodity: 2001 to 2015", line = 2, cex.main = 1.5, adj = 0)
# graphics::title(xlab="Counties", line=7, cex.lab = 1)
# graphics::title(ylab="Insurance claim loss ($)", line=4, cex.lab = 1)

```

\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Inland Pacific Northwest agricultural insurance loss for apples, by year, 2001 to 2022.")}

# par(mar=c(12.5, 6.1, 4, 2.1), family = 'serif', mgp=c(4, 1, 0), las=0)
# barplot(palouse_sumloss_year$loss, names.arg = palouse_sumloss_year$year, cex.names = 1, las = 3, yaxt="n",  col = "lightblue")
# pts <- pretty(palouse_sumloss_year$loss / 1000000)
# pts2 <- pretty(palouse_sumloss_year$loss)
# axis(2, las = 1, at = pts2, cex.axis = 1, labels = paste(pts, "M", sep = ""))
# 
# #mtext(text = "Years",
# #      side = 1,#side 1 = bottom
# #      line = 8)
# 
# graphics::title(main="Apples insurance loss by year: 2001 to 2015", line = 2, cex.main = 1.5, adj = 0)
# graphics::title(xlab="Years", line=6, cex.lab = 1)
# graphics::title(ylab="Insurance claim loss ($)", line=4, cex.lab = 1)

```

\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Inland Pacific Northwest agricultural insurance loss for wheat, by damage cause, 2001 to 2022.")}

usePackage("plotly")
usePackage("gridExtra")

#pie chart for 2007-2015

options(scipen=999)

addNoAnswer <- function(x){
  if(is.factor(x)) return(factor(x, levels=c(levels(x), "No Answer")))
  return(x)
}

RMA_damage_IPNW_com <- RMA_damage_IPNW
RMA_damage_IPNW_com$commodity <- tolower(RMA_damage_IPNW_com$commodity)
RMA_damage_IPNW_com$commodity <- stri_trans_general(RMA_damage_IPNW_com$commodity, id = "Title")
RMA_damage_IPNW_com$commodity <- factor(RMA_damage_IPNW_com$commodity)

WHEAT <- subset(RMA_damage_IPNW_com, commodity == "Wheat")

RMA_damage_IPNW_com2 <- WHEAT[WHEAT$damagecause == c("Drought", "Decline in Price", "Heat", "Cold Wet Weather", "Cold Winter", "Frost", "Hail", "Freeze", "Excess Moisture/Precipitation/Rain", "Excess Moisture/Precip/Rain"),]
  
#WHEAT <- subset(RMA_damage_IPNW_com2, commodity == "Wheat")

palouse_sumloss_1989_2015 <- aggregate(WHEAT$loss, list(WHEAT$damagecause), FUN = "sum")
colnames(palouse_sumloss_1989_2015) <- c("damagecause", "loss")

palouse_sumloss_1989_2015 <- palouse_sumloss_1989_2015[-1,]

palouse_sumloss_2007_2015_aggregate <- palouse_sumloss_1989_2015[order(palouse_sumloss_1989_2015$loss),] 
X <- palouse_sumloss_2007_2015_aggregate[order(palouse_sumloss_2007_2015_aggregate$loss),] 
#colnames(X) <- c("damagecause", "loss")
#summary_dataset <- X %>% filter(loss < 1500000) 
#summary_dataset2 <- X %>% filter(loss > 1500000) 
#colnames(summary_dataset) <- c("damagecause", "loss")
#Y <- sum(summary_dataset$loss)
#YY <- c("Other", Y)
#YY <- data.frame(t(YY))
#colnames(YY) <- c("damagecause", "loss")
#YYY <- rbind(summary_dataset2, YY)
#factor(YYY)
YYY <- X

moisture_combined <- YYY[YYY$damagecause == "Excess Moisture/Precipitation/Rain",]$loss + YYY[YYY$damagecause == "Excess Moisture/Precip/Rain",]$loss


YYY <- as.data.frame(lapply(YYY, addNoAnswer))
YYY$loss <- as.numeric(as.character(YYY$loss))
YYY[YYY$damagecause == "Excess Moisture/Precipitation/Rain",]$loss <- moisture_combined
YYY <- YYY[YYY$damagecause != "Excess Moisture/Precip/Rain",]
YYY$damagecause <- factor(YYY$damagecause)

levels(YYY$damagecause)[levels(YYY$damagecause)=="Excess Moisture/Precipitation/Rain"] <- "Excess Moisture"
YYY$damagecause <- factor(YYY$damagecause)

YYY1<- YYY[order(YYY$loss, decreasing = TRUE),] 
#YYY1 <- YYY1[ nrow(YYY1):1, ]
YYY1$damagecause <- factor(YYY1$damagecause)
#YYY1 <- YYY1[c(1:9),]

levels(YYY1[YYY1$damagecause == "Cold Wet Weather",]) <- "Cold Weather"
YYY1$damagecause <- factor(YYY1$damagecause)
YYY1 <- YYY1[c(1:9),]




par(mar=c(13.5, 6.1, 4, 2.1), family = 'serif', mgp=c(4, 1, 0), las=0)
barplot(YYY1$loss, names.arg = YYY1$damagecause, cex.names = 1.5, las = 3, cex.lab=1, yaxt="n",  col = "lightblue")
pts <- pretty(YYY$loss / 1000000)
pts2 <- pretty(YYY$loss)
axis(2, las = 1, at = pts2, cex.axis = 1, labels = paste(pts, "M", sep = ""), cex.axis=1.5)

#mtext(text = "Damage Cause",
#      side = 1,#side 1 = bottom
#      line = 8)

graphics::title(main="Wheat insurance loss by damage cause: 2001 to 2022", line = 2, cex.main = 1.5, adj = 0)
graphics::title(xlab="Damage Cause", line=11, cex.lab = 1.5)
graphics::title(ylab="Insurance claim loss ($)", line=5, cex.lab = 1.5)


#htmlTable(YYY2,
#          cgroup = c("2001-2015"),
#           caption = "IPNW region APPLES total insurance loss by damage cause, 2001-2015",
#          align = "c",
#          rnames = FALSE,
#          css.cell = "padding-left: .5em; padding-right: .5em;")

```

\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Inland Pacific Northwest stacked barplot of annual wheat agricultural insurance loss, by top damage causes, 2001 to 2022.")}

usePackage("maptools")

wheat_production <- read.csv("./data/production/wheat_production.csv")
wheat_production$legend <- c("Wheat Production ($)")

states <- readShapePoly('/tmp/aginsurance/states_conus/states_conus.shp', 
                                  proj4string=CRS
                                  ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
        projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

counties <- readShapePoly('/tmp/aginsurance/threestate_palouse/threestate_palouse.shp',
                     proj4string=CRS
                     ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")


#pal <- colorNumeric(palette = c("white", "orange", "darkorange", "red", "darkred"),
#                    domain = counties$NAME)
exte <- as.vector(extent(counties))

#label <- paste(sep = "<br/>", counties$NAME, round(counties$NAME, 0))
#markers <- data.frame(label)
#labs <- as.list(counties$NAME)

#counties <- counties[counties$NAME != "Kootenai",]
#counties <- counties[counties$NAME != "Clearwater",]


      lat_long <- coordinates(counties)


WHEAT <- subset(RMA_damage_IPNW_com, commodity == "Wheat")

WHEAT$damagecause <- as.factor(WHEAT$damagecause)


#levels(WHEAT$damagecause)[4] <- "Excessive Moisture"
WHEAT$damagecause <- factor(WHEAT$damagecause)

WHEAT_damage_top <- subset(WHEAT, damagecause == "Drought" | damagecause == "Heat" | damagecause == "Decline in Price" | damagecause == "Excessive Moisture")

WHEAT_damage_top$damagecause <- factor(WHEAT_damage_top$damagecause)
#aggregate(WHEAT_damage$loss, list(WHEAT_damage$damagecause), FUN = "sum")

usePackage("scales")

#ggplot(APPLES_damage_top, aes(fill=damagecause, y=loss, x=year)) + 
#    geom_bar( stat="identity") + theme_bw() + ggtitle("IPNW apples loss vs. year - top damage causes") + theme(axis.title.y = element_text(family = "Serif", size=16), axis.title.x = element_text(family = "Serif", size = 16), axis.text.x = element_text(size=rel(1.5), angle = 90, hjust = 1, family = "Serif")) + theme(plot.title = element_text(hjust = 0.5, family = "Serif"))  +  scale_x_discrete(limits=c(2001:2015)) + theme(legend.text=element_text(family = "Serif", size=14)) + theme(legend.title=element_text(family = "Serif", size=16, face = "bold")) + theme(plot.title = element_text(size=18)) + scale_y_continuous(label = unit_format(unit = "M", scale = 1e-6, digits = 0))

par(mar=c(11,8,6,2), mgp = c(5, 1, 0)) 

gg2 <- ggplot(WHEAT_damage_top) + 
    geom_bar(aes(fill=damagecause, y=loss, x=year), stat="identity") + theme_bw() + theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) + ggtitle("IPNW wheat loss vs. year - top damage causes") + theme(axis.title.y = element_text(size=20), axis.title.x = element_text(size = 20), axis.text.y = element_text(size=rel(1.2), hjust = 1), axis.text.x = element_text(size=rel(1.2), angle = 90, hjust = 1)) + theme(plot.title = element_text(size =20))  +  scale_x_discrete(name = "Year", limits=c(2001:2022)) + theme(legend.text=element_text(size=20)) + scale_fill_manual("legend", values = c("Drought" = "red", "Decline in Price" = "blue", "Heat" = "gray")) + theme(legend.title=element_text(size=20, face = "bold")) + theme(plot.title = element_text(size=20, face = "bold")) + scale_y_continuous(name = "Insurance claim loss ($)", labels = c("0 M", "500 M", "1 B", "1.5 B"), sec.axis = sec_axis(~ .*1, name="Wheat Production ($)", labels = c("0 M", "500 M", "1 B", "1.5 B"))) + theme(legend.position="bottom", legend.box = "vertical") + theme(text=element_text(size=20,  family="serif")) + geom_line(data = wheat_production, aes(x = year, y = utilization, colour = legend))+ scale_color_manual(values=c("black"), name="") +
  guides(color = guide_legend(nrow = 2)) 


print(gg2)

```

\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Inland Pacific Northwest map of wheat losses, 2001 to 2022.")}



usePackage("data.table")
usePackage("maptools")
usePackage("classInt")
usePackage("leaflet")
usePackage("dplyr")
usePackage("raster")
usePackage("htmltools")
usePackage("sp")
usePackage("latticeExtra")
library(RColorBrewer)

destfile <- "/data/states/states_conus.zip"
#download.file(URL, destfile)
outDir<-"/tmp/aginsurance/states_conus/"
unzip(destfile,exdir=outDir)

states <- readShapePoly('/tmp/aginsurance/states_conus/states_conus.shp',
                        proj4string=CRS
                        ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")


destfile <- "./data/counties/threestate_palouse.zip"
#download.file(URL, destfile)
outDir<-"/tmp/aginsurance/threestate_palouse/"
unzip(destfile,exdir=outDir)

counties_palouse <- readShapePoly('/tmp/aginsurance/threestate_palouse/threestate_palouse.shp',
                     proj4string=CRS
                     ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")


counties_palouse <- counties_palouse[counties_palouse$NAME != "Kootenai",]
counties_palouse <- counties_palouse[counties_palouse$NAME != "Clearwater",]

RMA_damage_IPNW_com <- RMA_damage_IPNW

#PNW damage cause as loadings using years as groups, JUST FOR WHEAT
RMA_damage_IPNW_com$commodity <- tolower(RMA_damage_IPNW_com$commodity)
RMA_damage_IPNW_com$commodity <- stri_trans_general(RMA_damage_IPNW_com$commodity, id = "Title")
RMA_damage_IPNW_com$commodity <- factor(RMA_damage_IPNW_com$commodity)

RMA_damage_IPNW_com <- RMA_damage_IPNW_com[RMA_damage_IPNW_com$commodity == c("Wheat"),]

RMA_damage_IPNW_com <- RMA_damage_IPNW_com[RMA_damage_IPNW_com$damagecause == c("Drought", "Heat"),]

RMA_damage_IPNW_com_map <- aggregate(RMA_damage_IPNW_com$loss, list(RMA_damage_IPNW_com$county), FUN = "sum")
colnames(RMA_damage_IPNW_com_map) <- c("NAME", "loss")


RMA_damage_IPNW_com_map2 <- merge(counties_palouse, RMA_damage_IPNW_com_map, by = "NAME")


 pal <- colorNumeric(colorRampPalette(c("lightyellow", "yellow", "orange", "red"))(24), na.color = "#ffffff",domain = eval(parse(text=paste("RMA_damage_IPNW_com_map2$", "loss", sep=""))))
 
 my.palette <- brewer.pal(n = 7, name = "OrRd")

 RMA_damage_IPNW_com_map2@data$id <- 1:nrow(RMA_damage_IPNW_com_map2@data)

 map.na <- list("SpatialPolygonsRescale", layout.north.arrow(type=1), offset = c(-121,47.5), scale = .5, col=1)

map.scale.1 <- list("SpatialPolygonsRescale", layout.scale.bar(),
                    offset = c(-121.5,47), scale = 1, col=1, fill =
                      c("transparent", "black"))
text0 <- list("sp.text", c(-121.5,47 + .12), "0", cex = 1)
text1 <- list("sp.text", c(-121.5 + 1.1, 47 + .12), "45mi", cex = .8)
text_WA <- list("sp.text", c(-121,46), "Washington", cex = .8)
text_ID <- list("sp.text", c(-115.8,46.8), "Idaho", cex = .8)
text_OR <- list("sp.text", c(-119,44.92), "Oregon", cex = .8)
text_MT <- list("sp.text", c(-115,47.5), "Montana", cex = .8)



  # Create list object for sp.layout (Got these functions here on stackoverflow, thank you owner)
    sp.label <- function(x, label) {list("sp.text", coordinates(x), label)}
    NUMB.sp.label <- function(x) {sp.label(x, as.vector(as.character(x@data$id)))}
    make.NUMB.sp.label <- function(x) {do.call("list", NUMB.sp.label(x))}
    
    map.layout.1 <-list(map.na,map.scale.1, text0, text1, text_WA, text_ID, text_OR, text_MT, states, make.NUMB.sp.label(RMA_damage_IPNW_com_map2))


spplot(RMA_damage_IPNW_com_map2, "loss", main = "Wheat loss: 2001 - 2022", sub = "Drought and Heat Damage Cause \n\n 1 Lewis: 2 Lincoln: 3 Garfield: 4 Columbia: 5 Walla Walla: 6 Umatilla: 7 Wallowa\n8 Nez Perce: 9 Douglas: 10 Wasco: 11 Latah: 12 Sherman: 13 Morrow: 14 Union\n15 Adams: 16 Whitman: 17 Asotin: 18 Grant: 19 Benewah: 20 Spokane\n21 Gilliam: 22 Benton: 23 Idaho: 24 Franklin", col.regions = my.palette, cuts =6, col = "gray", scales = list(draw = TRUE), sp.layout=list(map.layout.1), colorkey=list(at=seq(0, 140000000, 20000000), labels = c("0M", "20M", "40M", "60M", "80M", "100M", "120M", "140M")))

```


```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Inland Pacific Northwest agricultural insurance loss for apples, by county, 2001 to 2022.")}

usePackage("plotly")
usePackage("gridExtra")

#pie chart for 2007-2015

options(scipen=999)

addNoAnswer <- function(x){
  if(is.factor(x)) return(factor(x, levels=c(levels(x), "No Answer")))
  return(x)
}


RMA_damage_IPNW_com <- RMA_damage_IPNW

#PNW damage cause as loadings using years as groups, JUST FOR WHEAT
RMA_damage_IPNW_com$commodity <- tolower(RMA_damage_IPNW_com$commodity)
RMA_damage_IPNW_com$commodity <- stri_trans_general(RMA_damage_IPNW_com$commodity, id = "Title")
RMA_damage_IPNW_com$commodity <- factor(RMA_damage_IPNW_com$commodity)

APPLES <- RMA_damage_IPNW_com[RMA_damage_IPNW_com$commodity == c("Apples"),]


palouse_sumloss_1989_2015 <- aggregate(APPLES$loss, list(APPLES$damagecause), FUN = "sum")
colnames(palouse_sumloss_1989_2015) <- c("damagecause", "loss")

palouse_sumloss_2007_2015_aggregate <- palouse_sumloss_1989_2015[order(palouse_sumloss_1989_2015$loss),] 
X <- palouse_sumloss_2007_2015_aggregate[order(palouse_sumloss_2007_2015_aggregate$loss),] 
#colnames(X) <- c("damagecause", "loss")
summary_dataset <- X %>% filter(loss < 1500000) 
summary_dataset2 <- X %>% filter(loss > 1500000) 
colnames(summary_dataset) <- c("damagecause", "loss")
Y <- sum(summary_dataset$loss)
YY <- c("Other", Y)
YY <- data.frame(t(YY))
colnames(YY) <- c("damagecause", "loss")
YYY <- rbind(summary_dataset2, YY)
#factor(YYY)

YYY <- as.data.frame(lapply(YYY, addNoAnswer))
YYY$loss <- as.numeric(as.character(YYY$loss))
YYY1<- YYY[order(YYY$loss),] 
YYY1 <- YYY1[ nrow(YYY1):1, ]

#--year
palouse_sumloss_year <- aggregate(APPLES$loss, list(APPLES$year), FUN = "sum")
colnames(palouse_sumloss_year) <- c("year", "loss")

#--county

palouse_sumloss_county <- aggregate(APPLES$loss, list(APPLES$county), FUN = "sum")
colnames(palouse_sumloss_county) <- c("county", "loss")

p <- plot_ly(YYY, labels = ~damagecause, values = ~loss, type = 'pie',
             textposition = 'inside',
             textinfo = 'label+percent',
             insidetextfont = list(color = '#FFFFFF'),
             hoverinfo = 'text',
             text = ~paste('$', loss, ' million'),
             marker = list(colors = colors,
                           line = list(color = '#FFFFFF', width = 1)),
             #The 'pull' attribute can also be used to create space between the sectors
             showlegend = TRUE) %>%
  layout(title = 'Top Damage Causes for APPLES, Palouse Region: 1989-2015',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

p <- plot_ly(palouse_sumloss_year, labels = ~year, values = ~loss, type = 'pie',
             textposition = 'inside',
             textinfo = 'label+percent',
             insidetextfont = list(color = '#FFFFFF'),
             hoverinfo = 'text',
             text = ~paste('$', loss, ' million'),
             marker = list(colors = colors,
                           line = list(color = '#FFFFFF', width = 1)),
             #The 'pull' attribute can also be used to create space between the sectors
             showlegend = TRUE) %>%
  layout(title = 'Top Years for APPLES, Palouse Region: 1989-2015',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

p <- plot_ly(palouse_sumloss_county, labels = ~county, values = ~loss, type = 'pie',
             textposition = 'inside',
             textinfo = 'label+percent',
             insidetextfont = list(color = '#FFFFFF'),
             hoverinfo = 'text',
             text = ~paste('$', loss, ' million'),
             marker = list(colors = colors,
                           line = list(color = '#FFFFFF', width = 1)),
             #The 'pull' attribute can also be used to create space between the sectors
             showlegend = TRUE) %>%
  layout(title = 'Top Counties for APPLES, Palouse Region: 1989-2015',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

           
#p

YYY2 <- YYY1
YYY2$loss <- round(YYY2$loss, digits = 0)
YYY2$loss <- format(YYY2$loss,big.mark=",",scientific=FALSE)
YYY2$loss <- paste("$", YYY2$loss, sep="")
YYY2$loss <- gsub(" ", "", YYY2$loss, fixed = TRUE)

#--year
palouse_sumloss_year <- aggregate(APPLES$loss, list(APPLES$year), FUN = "sum")
colnames(palouse_sumloss_year) <- c("year", "loss")

#--county

palouse_sumloss_county <- aggregate(APPLES$loss, list(APPLES$county), FUN = "sum")
colnames(palouse_sumloss_county) <- c("county", "loss")

```


```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Inland Pacific Northwest agricultural insurance loss for apples, by damage cause, 2001 to 2022.")}
YYY1$damagecause <- factor(YYY1$damagecause)

#levels(YYY1$damagecause)[3] <- "Moisture"
#levels(YYY1$damagecause)[11] <- "Wind"

YYY1 <- YYY1[1:6,]

#YYY1$damagecause[8] <- "Excessive Moisture"
#YYY1$damagecause[11] <- "Snow/Lightning"

par(mar=c(13, 6.1, 4, 2.1), family = 'serif', mgp=c(4, 1, 0), las=0)
barplot(YYY1$loss, names.arg = YYY1$damagecause, cex.names = 1.5, las = 3, yaxt="n",  col = "lightblue")
pts <- pretty(YYY$loss / 1000000)
pts2 <- pretty(YYY$loss)
axis(2, las = 1, at = pts2, cex.axis = 1, labels = paste(pts, "M", sep = ""), cex.axis=1.5)

#mtext(text = "Damage Cause",
#      side = 1,#side 1 = bottom
#      line = 8)

graphics::title(main="Apples insurance loss by damage cause: 2001 to 2022", line = 2, cex.main = 1.5, adj = 0)
graphics::title(xlab="Damage Cause", line=11.5, cex.lab = 1.5)
graphics::title(ylab="Insurance claim loss ($)", line=4, cex.lab = 1.5)


#htmlTable(YYY2,
#          cgroup = c("2001-2015"),
#           caption = "IPNW region APPLES total insurance loss by damage cause, 2001-2015",
#          align = "c",
#          rnames = FALSE,
#          css.cell = "padding-left: .5em; padding-right: .5em;")

```

\newpage


```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Inland Pacific Northwest stacked barplot of annual apples agricultural insurance loss, by top damage causes, 2001 to 2022.")}

usePackage("maptools")

apple_production <- read.csv("./data/production/apple_production.csv")
apple_production$legend <- c("Apple Production ($)")

states <- readShapePoly('/tmp/aginsurance/states_conus/states_conus.shp', 
                                  proj4string=CRS
                                  ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
        projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

counties <- readShapePoly('/tmp/aginsurance/threestate_palouse/threestate_palouse.shp',
                     proj4string=CRS
                     ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")


#pal <- colorNumeric(palette = c("white", "orange", "darkorange", "red", "darkred"),
#                    domain = counties$NAME)
exte <- as.vector(extent(counties))

#label <- paste(sep = "<br/>", counties$NAME, round(counties$NAME, 0))
#markers <- data.frame(label)
#labs <- as.list(counties$NAME)

counties <- counties[counties$NAME != "Kootenai",]
counties <- counties[counties$NAME != "Clearwater",]


      lat_long <- coordinates(counties)


APPLES <- subset(RMA_damage_IPNW_com, commodity == "Apples")

APPLES$damagecause <- as.factor(APPLES$damagecause)

levels(APPLES$damagecause)[4] <- "Excessive Moisture"
APPLES$damagecause <- factor(APPLES$damagecause)

APPLES_damage <- subset(APPLES, damagecause == "Hail" | damagecause == "Frost" | damagecause == "Freeze" | damagecause == "Cold Wet Weather" | damagecause == "Wind/Excess Wind" | damagecause == "Heat")

APPLES_damage$damagecause <- factor(APPLES_damage$damagecause)
#aggregate(WHEAT_damage$loss, list(WHEAT_damage$damagecause), FUN = "sum")

usePackage("scales")

APPLES_damage_top <- subset(APPLES_damage, damagecause == "Hail" | damagecause == "Frost" | damagecause == "Freeze" | damagecause == "Cold Wet Weather" | damagecause == "Heat" | damagecause == "Wind/Excess Wind")
APPLES_damage_top <- subset(APPLES_damage_top, year >= 2001)


APPLES_damage_top_agg_hailfrostfreeze <- aggregate(APPLES_damage_top$loss, by = list(APPLES_damage_top$county), FUN = "sum")
colnames(APPLES_damage_top_agg_hailfrostfreeze) <- c("NAME", "loss")

county_apples_hailfrostfreeze <- merge(counties, APPLES_damage_top_agg_hailfrostfreeze, by = "NAME")

#2013

APPLES_damage_top2013 <- subset(APPLES_damage, damagecause == "Hail" | damagecause == "Frost" | damagecause == "Freeze")
APPLES_damage_top2013 <- subset(APPLES_damage_top2013, year == 2013)



APPLES_damage_top_agg_hailfrostfreeze2013 <- aggregate(APPLES_damage_top2013$loss, by = list(APPLES_damage_top2013$county), FUN = "sum")
colnames(APPLES_damage_top_agg_hailfrostfreeze2013) <- c("NAME", "loss")

county_apples_hailfrostfreeze2013 <- merge(counties, APPLES_damage_top_agg_hailfrostfreeze2013, by = "NAME")

#2006

APPLES_damage_top2006 <- subset(APPLES_damage, damagecause == "Hail" | damagecause == "Frost" | damagecause == "Freeze")
APPLES_damage_top2006 <- subset(APPLES_damage_top2006, year == 2006)



APPLES_damage_top_agg_hailfrostfreeze2006 <- aggregate(APPLES_damage_top2006$loss, by = list(APPLES_damage_top2006$county), FUN = "sum")
colnames(APPLES_damage_top_agg_hailfrostfreeze2006) <- c("NAME", "loss")

county_apples_hailfrostfreeze2006 <- merge(counties, APPLES_damage_top_agg_hailfrostfreeze2006, by = "NAME")


#how many claims of hail in walla walla in 2006?  2!

apples_hailfrostfreeze_count2006 <- subset(RMA_damage_IPNW, commodity == "APPLES")
apples_hailfrostfreeze_count2006 <- subset(apples_hailfrostfreeze_count2006, year == "2006")
apples_hailfrostfreeze_count2006 <- subset(apples_hailfrostfreeze_count2006, damagecause == "Hail" | damagecause == "Frost" | damagecause == "Freeze")


apples_hailfrostfreeze_count2006_sum <- aggregate(apples_hailfrostfreeze_count2006$count, by=list(apples_hailfrostfreeze_count2006$county), FUN = "sum")
colnames(apples_hailfrostfreeze_count2006_sum) <- c("county", "count")



#2015

APPLES_damage_top2015 <- subset(APPLES_damage, damagecause == "Hail" | damagecause == "Frost" | damagecause == "Freeze")
APPLES_damage_top2015 <- subset(APPLES_damage_top2015, year == 2015)



APPLES_damage_top_agg_hailfrostfreeze2015 <- aggregate(APPLES_damage_top2015$loss, by = list(APPLES_damage_top2015$county), FUN = "sum")
colnames(APPLES_damage_top_agg_hailfrostfreeze2015) <- c("NAME", "loss")

county_apples_hailfrostfreeze2015 <- merge(counties, APPLES_damage_top_agg_hailfrostfreeze2015, by = "NAME")


#2012

APPLES_damage_top2012 <- subset(APPLES_damage, damagecause == "Hail" | damagecause == "Frost" | damagecause == "Freeze")
APPLES_damage_top2012 <- subset(APPLES_damage_top2012, year == 2012)



APPLES_damage_top_agg_hailfrostfreeze2012 <- aggregate(APPLES_damage_top2012$loss, by = list(APPLES_damage_top2012$county), FUN = "sum")
colnames(APPLES_damage_top_agg_hailfrostfreeze2012) <- c("NAME", "loss")

county_apples_hailfrostfreeze2012 <- merge(counties, APPLES_damage_top_agg_hailfrostfreeze2012, by = "NAME")



#ggplot(APPLES_damage_top, aes(fill=damagecause, y=loss, x=year)) + 
#    geom_bar( stat="identity") + theme_bw() + ggtitle("IPNW apples loss vs. year - top damage causes") + theme(axis.title.y = element_text(family = "Serif", size=16), axis.title.x = element_text(family = "Serif", size = 16), axis.text.x = element_text(size=rel(1.5), angle = 90, hjust = 1, family = "Serif")) + theme(plot.title = element_text(hjust = 0.5, family = "Serif"))  +  scale_x_discrete(limits=c(2001:2015)) + theme(legend.text=element_text(family = "Serif", size=14)) + theme(legend.title=element_text(family = "Serif", size=16, face = "bold")) + theme(plot.title = element_text(size=18)) + scale_y_continuous(label = unit_format(unit = "M", scale = 1e-6, digits = 0))

par(mar=c(11,8,6,2), mgp = c(5, 1, 0)) 

# gg2 <- ggplot(APPLES_damage_top, aes(fill=damagecause, y=loss, x=year)) + 
#     geom_bar( stat="identity") + theme_bw() + theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) + ggtitle("IPNW apples loss vs. year - top damage causes") + theme(axis.title.y = element_text(size=20), axis.title.x = element_text(size = 20), axis.text.y = element_text(size=rel(1.2), hjust = 1), axis.text.x = element_text(size=rel(1.2), angle = 90, hjust = 1)) + theme(plot.title = element_text(size =20))  +  scale_x_discrete(name = "Year", limits=c(2001:2022)) + theme(legend.text=element_text(size=20)) + scale_fill_manual("legend", values = c("Hail" = "red", "Frost" = "blue", "Freeze" = "gray", "Cold Wet Weather" = "green", "Heat" = "magenta", "Wind/Excess Wind" = "purple")) + theme(legend.title=element_text(size=20, face = "bold")) + theme(plot.title = element_text(size=20, face = "bold")) + scale_y_continuous(name = "Insurance claim loss ($)", label = unit_format(unit = "M", scale = 1e-6, digits = 0)) + theme(legend.position="bottom") +
#   theme(text=element_text(size=20,  family="serif"))
# 
# print(gg2)


#--new


gg2 <- ggplot(APPLES_damage_top) + 
    geom_bar(aes(fill=damagecause, y=loss, x=year), stat="identity") + theme_bw() + theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) + ggtitle("IPNW apple loss vs. year - top damage causes") + theme(axis.title.y = element_text(size=20), axis.title.x = element_text(size = 20), axis.text.y = element_text(size=rel(1.2), hjust = 1), axis.text.x = element_text(size=rel(1.2), angle = 90, hjust = 1)) + theme(plot.title = element_text(size =20))  +  scale_x_discrete(name = "Year", limits=c(2001:2022)) + theme(legend.text=element_text(size=20)) + scale_fill_manual("legend", values = c("Hail" = "red", "Frost" = "blue", "Freeze" = "gray", "Cold Wet Weather" = "green", "Heat" = "magenta", "Wind/Excess Wind" = "purple")) + theme(legend.title=element_text(size=20, face = "bold")) + theme(plot.title = element_text(size=20, face = "bold")) + scale_y_continuous(name = "Insurance claim loss ($)", label = unit_format(unit = "M", scale = 1e-6, digits = 0), sec.axis = sec_axis(~ .*15, name="Apple Production ($)", labels = c("0 M", "500 M", "1 B", "1.5 B", "2.0 B", "2.5 B"))) + theme(legend.position="bottom", legend.box = "vertical") + theme(text=element_text(size=20,  family="serif")) + geom_line(data = apple_production, aes(x = year, y = utilization/15, colour = legend))+ scale_color_manual(values=c("black"), name="") +
  guides(color = guide_legend(nrow = 2)) 

print(gg2)

```
\newpage
<br></br>


```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Inland Pacific Northwest map of apple losses, 2001 to 2022.")}



usePackage("data.table")
usePackage("maptools")
usePackage("classInt")
usePackage("leaflet")
usePackage("dplyr")
usePackage("raster")
usePackage("htmltools")
usePackage("sp")
usePackage("latticeExtra")
library(RColorBrewer)

destfile <- "/data/states/states_conus.zip"
#download.file(URL, destfile)
outDir<-"/tmp/aginsurance/states_conus/"
unzip(destfile,exdir=outDir)

states <- readShapePoly('/tmp/aginsurance/states_conus/states_conus.shp',
                        proj4string=CRS
                        ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")


destfile <- "./data/counties/threestate_palouse.zip"
#download.file(URL, destfile)
outDir<-"/tmp/aginsurance/threestate_palouse/"
unzip(destfile,exdir=outDir)

counties_palouse <- readShapePoly('/tmp/aginsurance/threestate_palouse/threestate_palouse.shp',
                     proj4string=CRS
                     ("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
projection = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")


counties_palouse <- counties_palouse[counties_palouse$NAME != "Kootenai",]
counties_palouse <- counties_palouse[counties_palouse$NAME != "Clearwater",]

RMA_damage_IPNW_com <- RMA_damage_IPNW

#PNW damage cause as loadings using years as groups, JUST FOR WHEAT
RMA_damage_IPNW_com$commodity <- tolower(RMA_damage_IPNW_com$commodity)
RMA_damage_IPNW_com$commodity <- stri_trans_general(RMA_damage_IPNW_com$commodity, id = "Title")
RMA_damage_IPNW_com$commodity <- factor(RMA_damage_IPNW_com$commodity)

RMA_damage_IPNW_com <- RMA_damage_IPNW_com[RMA_damage_IPNW_com$commodity == c("Apples"),]

RMA_damage_IPNW_com <- RMA_damage_IPNW_com[RMA_damage_IPNW_com$damagecause == c("Freeze", "Hail", "Frost"),]

RMA_damage_IPNW_com_map <- aggregate(RMA_damage_IPNW_com$loss, list(RMA_damage_IPNW_com$county), FUN = "sum")
colnames(RMA_damage_IPNW_com_map) <- c("NAME", "loss")


RMA_damage_IPNW_com_map2 <- merge(counties_palouse, RMA_damage_IPNW_com_map, by = "NAME")


 pal <- colorNumeric(colorRampPalette(c("lightyellow", "yellow", "orange", "red"))(24), na.color = "#ffffff",domain = eval(parse(text=paste("RMA_damage_IPNW_com_map2$", "loss", sep=""))))
 
 my.palette <- brewer.pal(n = 7, name = "OrRd")

 RMA_damage_IPNW_com_map2@data$id <- 1:nrow(RMA_damage_IPNW_com_map2@data)

 map.na <- list("SpatialPolygonsRescale", layout.north.arrow(type=1), offset = c(-121,47.5), scale = .5, col=1)

map.scale.1 <- list("SpatialPolygonsRescale", layout.scale.bar(),
                    offset = c(-121.5,47), scale = 1, col=1, fill =
                      c("transparent", "black"))
text0 <- list("sp.text", c(-121.5,47 + .12), "0", cex = 1)
text1 <- list("sp.text", c(-121.5 + 1.1, 47 + .12), "45mi", cex = .8)
text_WA <- list("sp.text", c(-121,46), "Washington", cex = .8)
text_ID <- list("sp.text", c(-115.8,46.8), "Idaho", cex = .8)
text_OR <- list("sp.text", c(-119,44.92), "Oregon", cex = .8)
text_MT <- list("sp.text", c(-115,47.5), "Montana", cex = .8)



  # Create list object for sp.layout (Got these functions here on stackoverflow, thank you owner)
    sp.label <- function(x, label) {list("sp.text", coordinates(x), label)}
    NUMB.sp.label <- function(x) {sp.label(x, as.vector(as.character(x@data$id)))}
    make.NUMB.sp.label <- function(x) {do.call("list", NUMB.sp.label(x))}
    
    map.layout.1 <-list(map.na,map.scale.1, text0, text1, text_WA, text_ID, text_OR, text_MT, states, make.NUMB.sp.label(RMA_damage_IPNW_com_map2))


spplot(RMA_damage_IPNW_com_map2, "loss", main = "Apples loss: 2001 - 2022", sub = "Freeze, Frost, Hail Damage Cause \n\n 1 Lewis: 2 Lincoln: 3 Garfield: 4 Columbia: 5 Walla Walla: 6 Umatilla: 7 Wallowa\n8 Nez Perce: 9 Douglas: 10 Wasco: 11 Latah: 12 Sherman: 13 Morrow: 14 Union\n15 Adams: 16 Whitman: 17 Asotin: 18 Grant: 19 Benewah: 20 Spokane\n21 Gilliam: 22 Benton: 23 Idaho: 24 Franklin", col.regions = my.palette, cuts =6, col = "gray", scales = list(draw = TRUE), sp.layout=list(map.layout.1), colorkey=list(at=seq(0, 30000000, 5000000), labels = c("0M", "5M", "10M", "15M", "20M", "25M", "30M")))


#spplot(RMA_damage_IPNW_com_map2, "loss", col.regions = my.palette, cuts =6, col = "gray", scales = list(draw = TRUE), sp.layout=list(map.layout.1))

```


\newpage
<br></br>



```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of insurance loss for the entire PNW, for wheat by county, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used.")}

#PNW wheat

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2") 

RMA_damage_PNW_wheat <- subset(RMA_damage_PNW_com, commodity == "Wheat")

RMA_damage_PNW_transformed_wheat <- dcast(RMA_damage_PNW_wheat, county + state ~ damagecause, value.var = c("loss"), sum)
RMA_damage_PNW_transformed_wheat$statecounty <- paste(RMA_damage_PNW_transformed_wheat$county, "_", RMA_damage_PNW_transformed_wheat$state, sep="")
rownames(RMA_damage_PNW_transformed_wheat) <- RMA_damage_PNW_transformed_wheat$statecounty
#RMA_damage_PNW_exogenous_wheat <- RMA_damage_PNW_transformed_wheat[,3:27]
RMA_damage_PNW_exogenous_wheat <- RMA_damage_PNW_transformed_wheat

#ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_month), data = RMA_damage_PNW_transformed_month, colour = 'month', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 3, legend = FALSE) + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_month)) + theme(legend.position = "none")

#PNW wheat

#RMA_damage_PNW_exogenous_wheat <- cbind(RMA_damage_PNW_exogenous_wheat[,4:5], RMA_damage_PNW_exogenous_wheat[8], RMA_damage_PNW_exogenous_wheat[10], RMA_damage_PNW_exogenous_wheat[,13:16], RMA_damage_PNW_exogenous_wheat[18:19])

RMA_damage_PNW_exogenous_wheat <- RMA_damage_PNW_exogenous_wheat[3:12]

fit <- princomp(RMA_damage_PNW_exogenous_wheat, cor=TRUE)
#ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_wheat, scale = TRUE, center = TRUE), data = RMA_damage_PNW_transformed_wheat, colour = 'county', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 3, legend = FALSE) + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_wheat)) + theme(legend.position = "none")

plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_wheat, center = TRUE, scale = TRUE),  data = RMA_damage_PNW_transformed_wheat, loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, colour="state", label = FALSE, label.size = 2, legend = FALSE,  loadings.label.repel=TRUE)  + theme_bw()   + ggrepel::geom_text_repel(label=rownames(RMA_damage_PNW_exogenous_wheat), size = 3) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: PNW wheat insurance loss by county with \ndamage cause loadings ") + theme(plot.title = element_text(size =12, family = "Serif")) + theme(axis.title.y = element_text(family = "Serif", size=12), axis.title.x = element_text(family = "Serif", size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1, family = "Serif"), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif"))


#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot: insurance loss for the entire PNW, for wheat by county, with \ndamage cause as the factor loadings.", x = "principal components", y = "% explained variance")

grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))

```
\newpage
<br></br>

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of insurance loss for the entire PNW, for apples by county, with damage cause as the factor loadings.  Bottom panel: Scree plot. Data from 2001 is 2022 is used.")}

#PNW apples

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2")
RMA_damage_PNW_apples <- subset(RMA_damage_PNW_com, commodity == "Apples")

RMA_damage_PNW_transformed_apples <- dcast(RMA_damage_PNW_apples, county + state ~ damagecause, value.var = c("loss"), sum)
RMA_damage_PNW_transformed_apples$statecounty <- paste(RMA_damage_PNW_transformed_apples$county, "_", RMA_damage_PNW_transformed_apples$state, sep="")
rownames(RMA_damage_PNW_transformed_apples) <- RMA_damage_PNW_transformed_apples$statecounty
RMA_damage_PNW_exogenous_apples <- RMA_damage_PNW_transformed_apples[,3:10]

#RMA_damage_PNW_exogenous_apples <- cbind(RMA_damage_PNW_exogenous_apples[,1:4], RMA_damage_PNW_exogenous_apples[,7:10], RMA_damage_PNW_exogenous_apples[,13:14])

fit <- princomp(RMA_damage_PNW_exogenous_apples, cor=TRUE)
#PNW apples

#ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_apples), data = RMA_damage_PNW_transformed_apples, colour = 'county', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 3, legend = FALSE) + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_apples)) + theme(legend.position = "none")

plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_apples, center = TRUE, scale = TRUE),  data = RMA_damage_PNW_transformed_apples, colour = 'state', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 2, legend = FALSE, loadings.label.repel=TRUE)  + theme_bw()   + ggrepel::geom_text_repel(label=rownames(RMA_damage_PNW_exogenous_apples)) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: PNW apples insurance loss by county with \ndamage cause loadings ") + theme(plot.title = element_text(size =12, family = "Serif")) + theme(axis.title.y = element_text(family = "Serif", size=12), axis.title.x = element_text(family = "Serif", size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1, family = "Serif"), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif"))


#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot: insurance loss for the entire PNW, for apples by county, with \n damage cause as the factor loadings.", x = "principal components", y = "% explained variance")

grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))

```

\newpage
<br></br>


```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of insurance loss for the entire PNW, for barley by county, with damage cause as the factor loadings.  Bottom panel: Scree plot. Data from 2001 is 2022 is used.")}

#PNW barley

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2")

RMA_damage_PNW_barley <- subset(RMA_damage_PNW_com, commodity == "Barley")

RMA_damage_PNW_transformed_barley <- dcast(RMA_damage_PNW_barley, county + state ~ damagecause, value.var = c("loss"), sum)
RMA_damage_PNW_transformed_barley$statecounty <- paste(RMA_damage_PNW_transformed_barley$county, "_", RMA_damage_PNW_transformed_barley$state, sep="")
rownames(RMA_damage_PNW_transformed_barley) <- RMA_damage_PNW_transformed_barley$statecounty
#RMA_damage_PNW_exogenous_barley <- RMA_damage_PNW_transformed_barley[,3:23]
RMA_damage_PNW_exogenous_barley <- RMA_damage_PNW_transformed_barley[3:12]

#RMA_damage_PNW_exogenous_barley <- cbind(RMA_damage_PNW_exogenous_barley[,3:4], RMA_damage_PNW_exogenous_barley[7], RMA_damage_PNW_exogenous_barley[9], RMA_damage_PNW_exogenous_barley[,12:15], RMA_damage_PNW_exogenous_barley[17:18])

fit <- princomp(RMA_damage_PNW_exogenous_barley, cor=TRUE)

#PNW barley

#ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_barley), data = RMA_damage_PNW_transformed_barley, colour = 'county', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 3, legend = FALSE) + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_barley)) + theme(legend.position = "none")

plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_barley, center = TRUE, scale = TRUE), data = RMA_damage_PNW_transformed_barley, colour = 'state', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 2, legend = FALSE, loadings.label.repel=TRUE)  + theme_bw()   + ggrepel::geom_text_repel(label=rownames(RMA_damage_PNW_exogenous_barley)) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: PNW barley insurance loss by county with \ndamage cause loadings ") + theme(plot.title = element_text(size =12, family = "Serif")) + theme(axis.title.y = element_text(family = "Serif", size=12), axis.title.x = element_text(family = "Serif", size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1, family = "Serif"), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif"))


#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot: insurance loss for the entire PNW, for barley by county, with \ndamage cause as the factor loadings.", x = "principal components", y = "% explained variance")

grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))

```



\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of insurance loss for the inland PNW, for all commodities by county, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used.")}

#all PNW by damage

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2")
usePackage("gridExtra")


RMA_damage_PNW_transformed <- dcast(RMA_damage_PNW_com, county + state ~ damagecause, value.var = c("loss"), sum)
RMA_damage_PNW_transformed$statecounty <- paste(RMA_damage_PNW_transformed$county, "_", RMA_damage_PNW_transformed$state, sep="")
rownames(RMA_damage_PNW_transformed) <- RMA_damage_PNW_transformed$statecounty
RMA_damage_PNW_exogenous <- RMA_damage_PNW_transformed[,3:12]
par(mar=c(8, 6.1, 4, 2.1), family = 'serif', mgp=c(4, 1, 0), las=0)

fit <- stats::princomp(RMA_damage_PNW_exogenous, cor=TRUE)
plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous, center = TRUE, scale = TRUE), data = RMA_damage_PNW_transformed, colour = 'state', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 2, legend = FALSE,loadings.label.repel=TRUE)  + theme_bw()   + ggrepel::geom_text_repel( label=rownames(RMA_damage_PNW_exogenous), size=3) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: IPNW insurance loss by county with \ndamage cause loadings: 2001 to 2022") + theme(plot.title = element_text(size =12, family = "Serif")) + theme(axis.title.y = element_text(family = "Serif", size=12), axis.title.x = element_text(family = "Serif", size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1, family = "Serif"), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif"))

par(mar=c(8, 6.1, 4, 2.1), family = 'serif', mgp=c(4, 1, 0), las=0)

#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot of insurance loss for the inland PNW, for all commodities by county, with \ndamage cause as the factor loadings.", x = "principal components", y = "% explained variance")

grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))

```





\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of principal components for insurance loss for the entire iPNW, for all commodities by month, with damage cause as the factor loadings.  Bottom panel: Scree plot Data from 2001 is 2022 is used.")}

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2")

#PNW damage cause as loadings using months as groups

#remove negatives related to lossperacre calc
RMA_PNW_noneg <- subset(RMA_IPNW, lossperacre != Inf)
RMA_PNW_noneg2 <- subset(RMA_PNW_noneg, lossperacre >=0 )
RMA_PNW_noneg3 <- subset(RMA_PNW_noneg2, month !="" )

RMA_damage_PNW_transformed_month <- dcast(RMA_PNW_noneg3_com, month ~ damagecause, value.var = c("loss"), mean)

is.nan.data.frame <- function(x)
do.call(cbind, lapply(x, is.nan))

RMA_damage_PNW_transformed_month[is.nan(RMA_damage_PNW_transformed_month)] <- 0

#RMA_damage_PNW_transformed_year$statecounty <- paste(RMA_damage_PNW_transformed_year$county, "_", RMA_damage_PNW_transformed_year$state, sep="")
rownames(RMA_damage_PNW_transformed_month) <- RMA_damage_PNW_transformed_month$month
RMA_damage_PNW_exogenous_month <- RMA_damage_PNW_transformed_month[,2:10]

#RMA_damage_PNW_exogenous_month <- cbind(RMA_damage_PNW_exogenous_month[,2:3], RMA_damage_PNW_exogenous_month[6], RMA_damage_PNW_exogenous_month[8], RMA_damage_PNW_exogenous_month[,11:14], RMA_damage_PNW_exogenous_month[16:17])


#ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_year, center = TRUE, scale = TRUE), data = RMA_damage_PNW_transformed_year, colour = 'year', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 3, legend = FALSE) + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_year)) + theme(legend.position = "none")

#PNW using month as group - with damage cause as loadings
fit <- princomp(RMA_damage_PNW_exogenous_month, cor=TRUE)
plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_month, center = TRUE, scale = TRUE),  data = RMA_damage_PNW_transformed_month, loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 2, legend = FALSE)  + theme_bw()   + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_month)) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: iPNW insurance loss by month with damage cause loadings ") + theme(plot.title = element_text(size =12)) + theme(axis.title.y = element_text(size=12), axis.title.x = element_text(size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif"))
#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot: iPNW insurance loss, for all commodities by month, with damage cause \n as the factor loadings. ", x = "principal components", y = "% explained variance")

grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))

```

\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of insurance loss for the entire PNW, for barley by county, with damage cause as the factor loadings.  Bottom panel: Scree plot. Data from 2001 is 2022 is used.")}

#PNW barley

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2")


RMA_IPNW_com <- RMA_IPNW

#PNW damage cause as loadings using years as groups, JUST FOR WHEAT
RMA_IPNW_com$commodity <- tolower(RMA_IPNW_com$commodity)
RMA_IPNW_com$commodity <- stri_trans_general(RMA_IPNW_com$commodity, id = "Title")
RMA_IPNW_com$commodity <- factor(RMA_IPNW_com$commodity)


RMA_IPNW_com <- RMA_IPNW_com[RMA_IPNW_com$damagecause == c("Drought", "Decline in Price", "Heat", "Freeze", "Frost", "Cold Wet Weather", "Hail", "Excess Moisture/Precipitation/Rain", "Cold Winter", "Plant Disease"),]

RMA_damage_PNW_barley <- subset(RMA_IPNW_com, commodity == "Barley")

RMA_damage_PNW_transformed_barley <- dcast(RMA_damage_PNW_barley, county + state ~ damagecause, value.var = c("loss"), sum)
RMA_damage_PNW_transformed_barley$statecounty <- paste(RMA_damage_PNW_transformed_barley$county, "_", RMA_damage_PNW_transformed_barley$state, sep="")
rownames(RMA_damage_PNW_transformed_barley) <- RMA_damage_PNW_transformed_barley$statecounty
RMA_damage_PNW_exogenous_barley <- RMA_damage_PNW_transformed_barley[,3:10]
#RMA_damage_PNW_exogenous_barley <- RMA_damage_PNW_transformed_barley

#RMA_damage_PNW_exogenous_barley <- cbind(RMA_damage_PNW_exogenous_barley[,3:4], RMA_damage_PNW_exogenous_barley[7], RMA_damage_PNW_exogenous_barley[9], RMA_damage_PNW_exogenous_barley[,12:15], RMA_damage_PNW_exogenous_barley[17:18])

fit <- princomp(RMA_damage_PNW_exogenous_barley, cor=TRUE)

#PNW barley

#ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_barley), data = RMA_damage_PNW_transformed_barley, colour = 'county', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 3, legend = FALSE) + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_barley)) + theme(legend.position = "none")

plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_barley, center = TRUE, scale = TRUE), data = RMA_damage_PNW_transformed_barley,  loadings = TRUE, loadings.label = TRUE, colour = "state", loadings.label.size  = 3, label = FALSE, label.size = 2, legend = FALSE,loadings.label.repel=TRUE)  + theme_bw()   + ggrepel::geom_text_repel(label=rownames(RMA_damage_PNW_exogenous_barley), size=3) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: iPNW barley insurance loss by county with \ndamage cause loadings ") + theme(plot.title = element_text(size =12, family = "Serif")) + theme(axis.title.y = element_text(family = "Serif", size=12), axis.title.x = element_text(family = "Serif", size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1, family = "Serif"), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif"))


#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot: insurance loss for the entire iPNW, for barley by county, with \ndamage cause as the factor loadings.", x = "principal components", y = "% explained variance")

grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))

```



\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of insurance loss for the entire PNW, for wheat by county, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used.")}

#PNW wheat

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2") 


RMA_IPNW_com <- RMA_IPNW

#PNW damage cause as loadings using years as groups, JUST FOR WHEAT
RMA_IPNW_com$commodity <- tolower(RMA_IPNW_com$commodity)
RMA_IPNW_com$commodity <- stri_trans_general(RMA_IPNW_com$commodity, id = "Title")
RMA_IPNW_com$commodity <- factor(RMA_IPNW_com$commodity)


RMA_IPNW_com <- RMA_IPNW_com[RMA_IPNW_com$damagecause == c("Drought", "Decline in Price", "Heat", "Freeze", "Frost", "Cold Wet Weather", "Hail", "Excess Moisture/Precipitation/Rain", "Cold Winter", "Plant Disease"),]

RMA_damage_PNW_wheat <- subset(RMA_IPNW_com, commodity == "Wheat")

RMA_damage_PNW_transformed_wheat <- dcast(RMA_damage_PNW_wheat, county + state ~ damagecause, value.var = c("loss"), sum)
RMA_damage_PNW_transformed_wheat$statecounty <- paste(RMA_damage_PNW_transformed_wheat$county, "_", RMA_damage_PNW_transformed_wheat$state, sep="")
rownames(RMA_damage_PNW_transformed_wheat) <- RMA_damage_PNW_transformed_wheat$statecounty
RMA_damage_PNW_exogenous_wheat <- RMA_damage_PNW_transformed_wheat[,3:10]
#RMA_damage_PNW_exogenous_wheat <- RMA_damage_PNW_transformed_wheat

#ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_month), data = RMA_damage_PNW_transformed_month, colour = 'month', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 3, legend = FALSE) + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_month)) + theme(legend.position = "none")

#PNW wheat

#RMA_damage_PNW_exogenous_wheat <- cbind(RMA_damage_PNW_exogenous_wheat[,4:5], RMA_damage_PNW_exogenous_wheat[8], RMA_damage_PNW_exogenous_wheat[10], RMA_damage_PNW_exogenous_wheat[,13:16], RMA_damage_PNW_exogenous_wheat[18:19])

fit <- princomp(RMA_damage_PNW_exogenous_wheat, cor=TRUE)
#ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_wheat, scale = TRUE, center = TRUE), data = RMA_damage_PNW_transformed_wheat, colour = 'county', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 3, legend = FALSE) + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_wheat)) + theme(legend.position = "none")

plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_wheat, center = TRUE, scale = TRUE), data = RMA_damage_PNW_transformed_wheat, loadings = TRUE, loadings.label = TRUE, colour="state", loadings.label.size  = 3, label = FALSE, label.size = 2, legend = FALSE)  + theme_bw()   + ggrepel::geom_text_repel( label=rownames(RMA_damage_PNW_exogenous_wheat), size=3) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: iPNW wheat insurance loss by county with \ndamage cause loadings ") + theme(plot.title = element_text(size =12, family = "Serif")) + theme(axis.title.y = element_text(family = "Serif", size=12), axis.title.x = element_text(family = "Serif", size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1, family = "Serif"), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif"))


#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot: insurance loss for the entire iPNW, for wheat by county, with \ndamage cause as the factor loadings.", x = "principal components", y = "% explained variance")

grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))

```


\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of insurance loss for the entire iPNW, for apples by county, with damage cause as the factor loadings.  Bottom panel: Scree plot. Data from 2001 is 2022 is used.")}

#PNW apples

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2")


RMA_IPNW_com <- RMA_IPNW

#PNW damage cause as loadings using years as groups, JUST FOR WHEAT
RMA_IPNW_com$commodity <- tolower(RMA_IPNW_com$commodity)
RMA_IPNW_com$commodity <- stri_trans_general(RMA_IPNW_com$commodity, id = "Title")
RMA_IPNW_com$commodity <- factor(RMA_IPNW_com$commodity)


RMA_IPNW_com <- RMA_IPNW_com[RMA_IPNW_com$damagecause == c("Drought", "Decline in Price", "Heat", "Freeze", "Frost", "Cold Wet Weather", "Hail", "Excess Moisture/Precipitation/Rain", "Cold Winter", "Plant Disease"),]

RMA_damage_PNW_apples <- subset(RMA_IPNW_com, commodity == "Apples")

RMA_damage_PNW_transformed_apples <- dcast(RMA_damage_PNW_apples, county + state ~ damagecause, value.var = c("loss"), sum)
RMA_damage_PNW_transformed_apples$statecounty <- paste(RMA_damage_PNW_transformed_apples$county, "_", RMA_damage_PNW_transformed_apples$state, sep="")
rownames(RMA_damage_PNW_transformed_apples) <- RMA_damage_PNW_transformed_apples$statecounty
RMA_damage_PNW_exogenous_apples <- RMA_damage_PNW_transformed_apples[,3:10]

#RMA_damage_PNW_exogenous_apples <- cbind(RMA_damage_PNW_exogenous_apples[,1:4], RMA_damage_PNW_exogenous_apples[,7:10])

fit <- princomp(RMA_damage_PNW_exogenous_apples, cor=TRUE)
#PNW apples

#ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_apples), data = RMA_damage_PNW_transformed_apples, colour = 'county', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 3, legend = FALSE) + geom_text(vjust=-1, label=rownames(RMA_damage_PNW_exogenous_apples)) + theme(legend.position = "none")

plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_apples, center = TRUE, scale = TRUE), data = RMA_damage_PNW_transformed_apples, loadings = TRUE, loadings.label = TRUE, colour="state", loadings.label.size  = 3, label = FALSE, label.size = 2, legend = FALSE, loadings.label.repel=TRUE)  + theme_bw()   + ggrepel::geom_text_repel(label=rownames(RMA_damage_PNW_exogenous_apples), size=3) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: iPNW apples insurance loss by county with \ndamage cause loadings ") + theme(plot.title = element_text(size =12, family = "Serif")) + theme(axis.title.y = element_text(family = "Serif", size=12), axis.title.x = element_text(family = "Serif", size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1, family = "Serif"), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif"))


#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot: insurance loss for the entire iPNW, for apples by county, with \n damage cause as the factor loadings.", x = "principal components", y = "% explained variance")

grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))

```

\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.cap=paste("Using the results of our PCA for wheat by county, with damage cause as the factor loadings, we generate a scree plot that shows the proportion of variance explained by the individual components.")}

#just the IPNW

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2")


RMA_damage_IPNW_com <- RMA_damage_IPNW

#PNW damage cause as loadings using years as groups, JUST FOR WHEAT
RMA_damage_IPNW_com$commodity <- tolower(RMA_damage_IPNW_com$commodity)
RMA_damage_IPNW_com$commodity <- stri_trans_general(RMA_damage_IPNW_com$commodity, id = "Title")
RMA_damage_IPNW_com$commodity <- factor(RMA_damage_IPNW_com$commodity)


RMA_damage_IPNW_com <- RMA_damage_IPNW_com[RMA_damage_IPNW_com$damagecause == c("Drought", "Decline in Price", "Heat", "Freeze", "Frost", "Cold Wet Weather", "Hail", "Excess Moisture/Precipitation/Rain", "Cold Winter", "Plant Disease"),]



RMA_damage_IPNW_com$log10loss <- log10(RMA_damage_IPNW_com$loss)


RMA_damage_IPNW_wheat <- subset(RMA_damage_IPNW_com, commodity == "Wheat")
RMA_damage_IPNW_transformed <- dcast(RMA_damage_IPNW_wheat, county + state ~ damagecause, value.var = c("loss"), sum)
RMA_damage_IPNW_transformed$statecounty <- paste(RMA_damage_IPNW_transformed$county, "_", RMA_damage_IPNW_transformed$state, sep="")
rownames(RMA_damage_IPNW_transformed) <- RMA_damage_IPNW_transformed$statecounty
RMA_damage_IPNW_exogenous <- RMA_damage_IPNW_transformed[,3:10]


#IPNW

#RMA_damage_IPNW_exogenous <- scale(RMA_damage_IPNW_exogenous, scale = TRUE, center = TRUE)
pca_RMA_damage_IPNW_exogenous <- prcomp(RMA_damage_IPNW_exogenous, scale = TRUE, center = TRUE)

#RMA_damage_IPNW_exogenous_factor <- cbind(RMA_damage_IPNW_exogenous[2:3], RMA_damage_IPNW_exogenous[5:6], RMA_damage_IPNW_exogenous[8], RMA_damage_IPNW_exogenous[11], RMA_damage_IPNW_exogenous[13:14], RMA_damage_IPNW_exogenous[16:17])


RMA_damage_IPNW_exogenous_factor <- RMA_damage_IPNW_exogenous
fit <- prcomp(RMA_damage_IPNW_exogenous_factor, cor=TRUE, center = TRUE, scale = TRUE)

comp <- data.frame(fit$x[,1:2])
# Plot
#plot(comp, pch=16, col=rgb(0,0,0,0.5))


wss <- (nrow(RMA_damage_IPNW_exogenous_factor)-1)*sum(apply(RMA_damage_IPNW_exogenous_factor,2,var))
for (i in 1:23) wss[i] <- sum(kmeans(RMA_damage_IPNW_exogenous_factor,
                                     centers=i)$withinss)
#plot(1:23, wss, type="b", xlab="Number of Clusters", yaxt = "n", 
#     ylab="Within groups sum of squares", main = "KMeans COUNTIES Elbow Plot")
#axis(2, las = 1, at = wss, cex.axis = 1, labels = paste(round(wss/100000000000000, 0), "M", sep = ""))

```
\newpage




``` {r, warning = FALSE, echo= FALSE, message = FALSE, fig.show = 'hide', results = 'hide'}

#``` {r, warning = FALSE, echo= FALSE, message = FALSE, results='hide',fig.keep='all', fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of insurance loss for the inland PNW, for all commodities by county, with damage cause as the factor loadings.  Bottom panel: Scree plot. Data from 2001 is 2022 is used")}

  #identify the optimum number of clusters using NbClust
  par(mar=c(1,1,1,1))
  usePackage("NbClust")
  usePackage("factoextra")
  
  clustering_distance <- c("euclidean")
 clust<-NbClust(fit$x[,1:3], diss=NULL, distance = clustering_distance, min.nc=2, max.nc=20, method = "kmeans", index = "all")


  #clust<-NbClust(damage_loss[,1:(length(waterscarcity)-4)], diss=NULL, distance = clustering_distance, min.nc=2, max.nc=20, method = "kmeans", index = "dindex") 
  
  #res$All.index
  #res$Best.nc
  #res$All.CriticalValues
  #res$Best.partition
  
  maxclusters <- length(unique(clust$Best.partition))
  km.res <- kmeans(fit$x[,1:2], maxclusters, nstart = 25)
  #pam.res <- pam(damage_loss, maxclusters)
  hc.cut <- hcut(fit$x[,1:2], k = maxclusters, hc_method = "ward.D2")

```
\newpage



``` {r, warning = FALSE, echo= FALSE, message = FALSE, fig.show = 'hide', results = 'hide'}

  #identify the optimum number of clusters using NbClust
  par(mar=c(1,1,1,1))
  usePackage("NbClust")
  usePackage("factoextra")
  
  clustering_distance <- c("euclidean")
 clust<-NbClust(fit$x[,1:3], diss=NULL, distance = clustering_distance, min.nc=2, max.nc=20, method = "kmeans", index = "all")
  par(mfrow=c(1,1))  
  
```
\newpage

```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, results = 'hide', fig.width = 8, fig.height = 8, fig.cap=paste("Hierarchical clustering dendrogram of Inland Pacific Northwest counties, using the first two principal components of the aforementioned dimensionality reduction exercise. This clustering method provides an alternative approach as compared to Kmeans clustering.")}

par(mar=c(8, 6.1, 4, 2.1), family = 'serif', mgp=c(4, 1, 0), las=0)

#plot(hc.cut, cex = 1, xlab = "Counties") # plot tree
#rect.hclust(hc.cut, k = maxclusters, border = 2:5) # add rectangle

usePackage("mclust")
d_clust <- Mclust(RMA_damage_IPNW_exogenous_factor, G=1:7, 
                  modelNames = mclust.options("emModelNames"))
#d_clust$BIC
#plot(d_clust, what = c("BIC"))


```




```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of insurance loss for the entire iPNW, for wheat by year, with damage cause as the factor loadings. Bottom panel: Scree plot. Data from 2001 is 2022 is used.")}

usePackage("ggplot2")
usePackage("ggfortify")
usePackage("reshape2")

RMA_IPNW_com <- RMA_IPNW

#PNW damage cause as loadings using years as groups, JUST FOR WHEAT
RMA_IPNW_com$commodity <- tolower(RMA_IPNW_com$commodity)
RMA_IPNW_com$commodity <- stri_trans_general(RMA_IPNW_com$commodity, id = "Title")
RMA_IPNW_com$commodity <- factor(RMA_IPNW_com$commodity)


RMA_IPNW_com <- RMA_IPNW_com[RMA_IPNW_com$damagecause == c("Drought", "Decline in Price", "Heat", "Freeze", "Frost", "Cold Wet Weather", "Hail", "Excess Moisture/Precipitation/Rain", "Cold Winter", "Plant Disease"),]

RMA_damage_PNW_wheat_year <- subset(RMA_IPNW_com, commodity == "Wheat")



RMA_damage_PNW_transformed_wheat_year <- dcast(RMA_damage_PNW_wheat_year, year ~ damagecause, value.var = c("loss"), sum)
#RMA_damage_PNW_transformed_year$statecounty <- paste(RMA_damage_PNW_transformed_year$county, "_", RMA_damage_PNW_transformed_year$state, sep="")
rownames(RMA_damage_PNW_transformed_wheat_year) <- RMA_damage_PNW_transformed_wheat_year$year
RMA_damage_PNW_exogenous_wheat_year <- RMA_damage_PNW_transformed_wheat_year
par(mar=c(8, 6.1, 4, 2.1), family = 'serif', mgp=c(4, 1, 0), las=0)

#PNW using year as group - with damage cause as loadings: JUST FOR WHEAT

#RMA_damage_PNW_exogenous_wheat_year_factor <- cbind(RMA_damage_PNW_exogenous_wheat_year[,3:4], RMA_damage_PNW_exogenous_wheat_year[7], RMA_damage_PNW_exogenous_wheat_year[9], RMA_damage_PNW_exogenous_wheat_year[,12:15], RMA_damage_PNW_exogenous_wheat_year[17:18])
fit <- princomp(RMA_damage_PNW_exogenous_wheat_year, cor=TRUE)
fit2 <- prcomp(RMA_damage_PNW_exogenous_wheat_year, cor=TRUE, center = TRUE, scale = TRUE)

comp <- data.frame(fit2$x[,1:2])


usePackage("RColorBrewer")
usePackage("scales")

k <- kmeans(comp, 2, nstart=25, iter.max=1000)

palette(alpha(brewer.pal(9,'Set2'), 1))
#plot(comp, col=k$clust, pch=2)


kclust <- data.frame(k$clust)
kclust$k.clust <- factor(kclust$k.clust)

plot1 <- ggplot2::autoplot(prcomp(RMA_damage_PNW_exogenous_wheat_year, center = TRUE, scale = TRUE),data = kclust, colour = 'k.clust',   loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 4, label = FALSE, label.size = 2, legend = FALSE,loadings.label.repel=TRUE)  +  theme_bw()   + ggrepel::geom_text_repel(label=rownames(RMA_damage_PNW_exogenous_wheat_year), segment.colour = NA, size=4) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: iPNW wheat insurance loss by year with \ndamage cause loadings ") + theme(plot.title = element_text(size =12, family = "Serif")) + theme(axis.title.y = element_text(family = "Serif", size=12), axis.title.x = element_text(family = "Serif", size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1, family = "Serif"), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif")) + ggforce::geom_mark_ellipse(aes(color = k.clust)) + theme(legend.position = 'bottom') + coord_equal(ylim = c(-.7, .8), xlim = c(-.7, .8))



#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot: insurance loss for the entire iPNW, for wheat by year, with \ndamage cause as the factor loadings.", x = "principal components", y = "% explained variance")

grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))


```



```{r message=FALSE, warning=FALSE, echo= FALSE, error=TRUE, fig.width = 8, fig.height = 8, fig.cap=paste("Top panel: biplot of insurance loss for the iPNW, for wheat by county, with damage cause as the factor loadings.  Uses a Kmeans technique for grouping. Bottom panel: Scree plot. Data from 2001 is 2022 is used.")}

usePackage("RColorBrewer")
usePackage("scales")

fit <- prcomp(RMA_damage_IPNW_exogenous_factor, cor=TRUE, center = TRUE, scale = TRUE)

comp <- data.frame(fit$x[,1:2])

k <- kmeans(comp, 2, nstart=25, iter.max=1000)

palette(alpha(brewer.pal(9,'Set2'), 1))
#plot(comp, col=k$clust, pch=2)


kclust <- data.frame(k$clust)
kclust$k.clust <- factor(kclust$k.clust)

# plot1 <- ggplot2::autoplot(prcomp(RMA_damage_IPNW_exogenous_factor, scale = TRUE, center = TRUE), frame = TRUE, frame.type = 'norm', data = kclust, colour = 'k.clust', loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3, label = FALSE, label.size = 3, legend = FALSE,loadings.label.repel=TRUE)  + theme_bw()   + ggrepel::geom_text_repel(label=rownames(RMA_damage_IPNW_exogenous), size=2) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: iPNW insurance loss\nby county with damage cause loadings + Kmeans clustering method") + theme(plot.title = element_text(size =12, family = "Serif")) + theme(axis.title.y = element_text(family = "Serif", size=12), axis.title.x = element_text(family = "Serif", size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1, family = "Serif"), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif"))  + theme(legend.position="bottom") 


plot1 <- ggplot2::autoplot(prcomp(RMA_damage_IPNW_exogenous_factor, center = TRUE, scale = TRUE),data = kclust, colour = 'k.clust',   loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 4, label = FALSE, label.size = 2, legend = FALSE,loadings.label.repel=TRUE)  +  theme_bw()   + ggrepel::geom_text_repel(label=rownames(RMA_damage_IPNW_exogenous_factor), segment.colour = NA, size=4) + theme(legend.position = "none") + ggtitle("Principal Components Analysis: iPNW insurance loss\nby county with damage cause loadings + Kmeans clustering method ") + theme(plot.title = element_text(size =12, family = "Serif")) + theme(axis.title.y = element_text(family = "Serif", size=12), axis.title.x = element_text(family = "Serif", size = 12), axis.text.x = element_text(size=rel(1.5), hjust = 1, family = "Serif"), axis.text.y = element_text(size=rel(1.5), hjust = 1, family = "Serif")) + ggforce::geom_mark_ellipse(aes(color = k.clust)) + theme(legend.position = 'bottom') + coord_equal(xlim = c(-.50, .8), ylim=c(-.70, .60))

#scree plot

std_dev <- fit$sdev

pr_var <- std_dev^2

#pr_var[1:10]

prop_varex <- pr_var/sum(pr_var)

prop_varex <- data.frame(prop_varex)

prop_varex$comp <- c(1:max(nrow(prop_varex)))

prop_varex$prop_varex <- as.numeric(prop_varex$prop_varex)

p <- ggplot(prop_varex, aes(as.factor(comp), prop_varex, group=1))
plot2 <- p + geom_point(shape = 21) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  xlab("comp") + theme_bw() +
stat_summary(fun.y=sum, geom="line") + labs(title = "Scree plot of insurance loss for the inland iPNW, for wheat by county, with \ndamage cause as the factor loadings.", x = "principal components", y = "% explained variance")

grid.arrange(plot1, plot2, nrow=2, heights=c(2,1))


```
